### 2013-02-26 paper/paper.tex
bootstrapping with the tt-ba paper

\documentclass[11pt,a4paper]{article}
%\usepackage{lrec2012}
%\usepackage{url}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=blue]{hyperref}
\usepackage{times}
\usepackage{multirow}
\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}
\usepackage{latexsym}
\usepackage[small,bf]{caption}
\usepackage{xltxtra}

\usepackage{fontspec}
\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
\newfontfeature{IPA}{+mgrk}
%\setromanfont[IPA]{FreeSerif}
\setromanfont[Scale=0.9]{Times New Roman}
\newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}

\newcommand{\email}[1]{\texttt{\href{mailto:#1}{#1}}}
\usepackage{natbib}
%\usepackage{natbib,natbibspacing}
\usepackage{setspace}
\usepackage{booktabs}

%\setlength{\bibspacing}{\baselineskip-0.4em}

\newenvironment{itemise}[1]{
        \begin{itemize}\setlength{\itemsep}{-0.3em}
		  %\setlength{\itemindent}{-1em}
		  %\setlength{\leftmargin}{-2em}
        \vspace{-0.6em}
        #1
}{
        \end{itemize}
        \vspace{-1pt}
}

%\setlength\titlebox{6.5cm}    % You can expand the title box if you
% really have to

\title{A prototype machine translation system for Tatar and Bashkir based on 
   free/open-source components}

\name{Ilnar Salimzyanov, Jonathan North Washington, and Francis M. Tyers}

\address{ Kazan Federal University, Indiana University, Universitat d'Alacant \\
          Kazan, Bloomington, Alacant\\
          Russia, IN 47405 (USA), E-03071 Spain\\
          \email{ilnar.salimzyan@gmail.com}, \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es} \\
}

\newcommand{\tag}[1]{{\small{\texttt{#1}}}}

\date{\today}

\abstract{
This paper presents a prototype bidirectional machine translation system between 
Tatar and Bashkir, two minority Turkic languages of Russia. While the 
system has low open-domain coverage, results are presented that suggest
that high accuracy may be obtained between these two closely-related
languages, on a par with similar systems.
 \\
\Keywords{Tatar, Bashkir, MT, Free software, Open-source}
}

\begin{document}
\spacing{1}
\maketitleabstract

\section{Introduction}

This paper presents a prototype shallow-transfer rule-based machine translation
system between Tatar and Bashkir, two closely-related minority Turkic 
languages of Russia.

The paper will be laid out as follows: Section~\ref{sec:lang}\ gives a brief description
of the two languages; Section~\ref{sec:prev}\ gives a short review of some previous
work in the area of Turkic--Turkic language translation; Section~\ref{sec:sys}\ 
describes the system and the tools used to construct it; Section~\ref{sec:eval}\ gives a very preliminary evaluation of the system; and finally Section~\ref{sec:conc}\ describes our aims for future work and some concluding remarks.

\section{Languages}
\label{sec:lang}

Tatar is a Turkic language spoken in and around Tatarstan by approximately 6 million people.  Bashkir (Bashqort) is a Turkic language spoken by about 1.5 million people in and around Bashqortostan.  Tatarstan and Bashqortostan are both republics within Russia.  Both languages are co-official with Russian in their respective republics.  The two languages belong to the same branch of the Kypchak group of Turkic languages.  As they are very close relatives, they share many innovations, but Bashkir has quite a few phonological innovations beyond those of Tatar (such as rounding harmony and desonorisation of high-sonority suffix-initial consonants; cf.\ \citet{washington10}) and the languages have a number of morphological differences (including different volitional participles).  The spoken languages share a high level of mutual intelligibility, but many of the inherent similarities are obscured by their fairly different orthographical systems along with the phonological and morphological differences between the languages.

Aside from native speaker intuition, we also consulted the Bashkir Grammar of M. G. Usmanova \citep{usmanova06}.

\section{Previous work}
\label{sec:prev}

Several previous works on making machine translation systems between 
Turkic languages exist, although to our knowledge none are publically available.

For systems between Turkish and other Turkic languages, there have been, 
for example, systems reported for Turkish-Crimean Tatar \citep{altintas01}, Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and Turkish-Turkmen \citep{tantug07}.

\section{System}
\label{sec:sys}

\begin{figure*}[htbp]
\begin{center}
 \includegraphics[width=0.8\textwidth]{architecture.pdf}
\end{center}
\caption{The pipeline architecture of the Apertium system.}
\label{fig:modules}
%\vspace{-1em}
\end{figure*}

The system is based on the Apertium machine translation 
platform \citep{apertium/2011}.\footnote{\url{http://www.apertium.org}} The 
platform was originally aimed at the Romance languages of the Iberian peninsula, but has also been adapted for 
other, more distantly related, language pairs.
The whole platform, both programs and data, are licensed under the Free Software Foundation's General Public 
Licence\footnote{\url{http://www.fsf.org/licensing/licenses/gpl.html}} (GPL) and all the software and data for the 
30 supported language pairs (and the other pairs being worked on) is available for download from the project 
website.

\subsection{Architecture of the system}

The Apertium translation engine consists of a Unix-style \emph{pipeline} or
\emph{assembly line} with the following modules (see Fig.~\ref{fig:modules}):  
\begin{itemise}
\item A \emph{deformatter} which encapsulates the format information
 in the input as \emph{superblanks} that will then be seen
 as blanks between words by the other modules.
\item A \emph{morphological analyser} which segments the text in
  surface forms (SF) (\emph{words}, or, where detected, multi-word lexical
  units or MWLUs) and for each, delivers one or more \emph{lexical
    forms} (LF) consisting of \emph{lemma}, \emph{lexical category} and
  morphological information. 
\item A \emph{morphological disambiguator} (constraint grammar) which chooses, using linguistic rules
  the most adequate sequence of morphological analyses for an ambiguous sentence. 
\item A \emph{lexical transfer} module which reads each SL LF 
  and delivers the corresponding target-language (TL) LF
  by looking it up in a bilingual dictionary encoded as an FST
  compiled from the corresponding XML file. The lexical transfer module may
  return more than one TL LF for a single SL LF.
\item A \emph{lexical selection} module which chooses, based on context 
  rules the most adequate translation of ambiguous source language LFs.
\item A \emph{structural transfer} module which
    performs local syntactic operations, is compiled from XML files containing rules that 
    associate an \emph{action} to each defined LF \emph{pattern}. Patterns are applied left-to-right, and the 
    longest matching pattern is always selected.
\item A \emph{morphological generator} which delivers a TL SF
 for each TL LF, by suitably inflecting it. 
\item A \emph{reformatter} which de-encapsulates any format
  information.
\end{itemise}

\subsection{Morphological transducers}

The morphological transducers are based on the Helsinki Finite State Toolkit \citep{hfst/2011}, a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis. It implements both the \textbf{lexc} formalism for defining lexicons, and the \textbf{twol} and \textbf{xfst} formalisms for modeling morphophonological rules. It also supports other finite state transducer formalisms such as \textbf{sfst}. This toolkit has been chosen as it -- or the equivalent XFST -- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006}, and is available under a free/open-source licence.

The morphologies of both languages are implemented in lexc, and the morphophonologies of both languages are implemented in twol.

Use of lexc allows for straightforward definition of different word classes and subclasses.  For example, Tatar (but not Bashkir) has two classes of verbs: one which take a harmonised high vowel in the infinitive (the default), and one which take a harmonised low vowel in the infinitive.  This was implemented in lexc with two similar continuation lexica for verbs: one pointing at a lexicon with an A-initial infinitive ending, and another pointing at a lexicon with an I-initial infinitive ending.

Use of twol allows for phonological processes present in the languages, like vowel harmony and desonorisation, to be implemented in a straightforward manner.  For example, in Tatar, the A and I archiphonemes found in the infinitive are harmonised to one of two vowels each, depending on the value of the preceding vowel; the basic form of this process can be implemented in one twol rule.

The same morphological description is used for both analysis and generation. To avoid overgeneration, any alternative forms are 
marked with one of two marks, {\tt {\small LR}} (only analyser) or {\tt {\small RL}} (only generator). Instead of the usual
compile/invert to compile the transducers, we compile twice, once the generator, without the {\tt {\small LR}} paths, and
then again the analyser without the {\tt {\small RL}} paths.

\subsection{Bilingual lexicon}
\begin{figure*}[htbp]
\begin{texttt}
    <e><p><l>\textbf{юнәлеш}<s n="n"/></l><r>\textbf{йүнәлеш}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{танау}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
\end{texttt}
\caption{Example entries from the bilingual transfer lexicon. Tatar is on the left, and Bashkir on the right}
\label{fig:bidix}
%\vspace{-1em}
\end{figure*}

The bilingual lexicon currently contains 2,834 stem to stem correspondences and was build by hand by a bilingual 
speaker of Tatar and Bashkir, translating a frequency list of the Russian National Corpus\footnote{\url{http://ruscorpora.ru/en/}} into both languages in a spreadsheet. This spreadsheet was then converted into the Apertium
XML dictionary format.

Entries consist largely of one-to-one stem-to-stem correspondences with part of speech, but also
include some entries with ambiguous translations (see e.g., Fig.~\ref{fig:bidix}).

\subsection{Disambiguation rules}

The system has a morphological disambiguation module in the form of a 
Constraint Grammar (CG) \citep{karlsson95}. The version of the formalism used is 
vislcg3.\footnote{\url{http://beta.visl.sdu.dk/constraint_grammar.html}}

The grammar currently has only four rules, but given the closeness of the languages, the 
majority of ambiguity may be passed through from one language to the other.

\subsection{Lexical selection rules}

Likewise, lexical selection is not a large problem between Tatar and Bashkir, but a 
number of rules can be written for ambiguous words; for example, the Tatar 
word \emph{борын} `nose (person), nose (ship)' can be translated into Bashkir 
as either \emph{танау} `nose (person)' or \emph{морон} `nose (ship)'. A lexical selection
rule chooses the translation \emph{танау} if the immediate context includes a proper 
name.

Another example is the word \emph{катлаулы} `layered'.  It is always translated to Bashkir 
as \emph{{\qipa ҡ}атмарлы}, except in the collocaton \emph{катлаулы мәсьәлә} `difficult matter/problem', which is translated as \emph{{\qipa ҡ}атлаулы мәсьәлә}.

%\subsection{Transfer rules}

\begin{table*}[htbp]
\centering
\begin{tabular}{ll}
%\hline
%{\bf Stage} & {\bf Representation} \\
\toprule
{\bf (Tatar) Input} & Һава бүген бик әйбәт, җылы гына. \\ 
\midrule
{\bf Mor. analysis} & \^{}Һава/һава\tag{<n>}\tag{<attr>}/һава\tag{<n>}\tag{<nom>\$} \^{}бүген/бүген\tag{<adv>\$}  \\
 ~             & \^{}бик/бик\tag{<adv>}/бик\tag{<n>}\tag{<attr>}/бик\tag{<n>}\tag{<nom>\$} \\ 
 ~             & \^{}әйбәт/әйбәт\tag{<adj>}/әйбәт\tag{<adj>}\tag{<subst>}\tag{<nom>\$}\^{},/,\tag{<cm>\$} \\ 
 ~             & \^{}җылы/җылы\tag{<n>}\tag{<attr>}/җылы\tag{<n>}\tag{<nom>}/җылы\tag{<adj>}/җылы\tag{<adj>}\tag{<subst>}\tag{<nom>\$} \\
 ~             & \^{}гына/гына\tag{<postadv>\$}\^{}./.\tag{<sent>\$} \\
\midrule
{\bf Mor. disambiguation}& \^{}Һава\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
 ~                  & \^{}җылы\tag{<adj>\$} \^{}гына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
\midrule
{\bf Lex. transfer} & \^{}Һава\tag{<n>}\tag{<nom>}/Һауа\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>}/бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>}/бик\tag{<adv>\$} \\
( + selection)                     & \^{}әйбәт\tag{<adj>}/әйбәт\tag{<adj>\$}\^{},\tag{<cm>}/,\tag{<cm>\$} \^{}җылы\tag{<adj>}/йылы\tag{<adj>\$} \\ 
~                     & \^{}гына\tag{<postadv>}/ғына\tag{<postadv>\$}\^{}.\tag{<sent>}/.\tag{<sent>\$}\\
\midrule
{\bf Struct. transfer}& \^{}Һауа\tag{<n>}\tag{<nom>\$} \^{}бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
~                    & \^{}йылы\tag{<adj>\$} \^{}ғына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
\midrule
{\bf Mor. generation} & Һауа бөгөн бик әйбәт, йылы ғына. \\
\bottomrule
\end{tabular}
 \caption{Translation process for the phrase \emph{Һава бүген бик әйбәт, җылы гына} `The weather today is very nice, it is very warm'.}
 %\vspace{-1em}
\end{table*}

\section{Evaluation}
\label{sec:eval}

Lexical coverage of the system is calculated over a freely available corpus of Bashkir, the Bashkir
Wikipedia,\footnote{\url{http://ba.wikipedia.org/}; {\tt bawiki-20111210-pages-articles.xml.bz2}} and over two freely available corpora of 
Tatar, the Tatar Wikipedia\footnote{\url{http://tt.wikipedia.org/}; {\tt ttwiki-20111215-pages-articles.xml.bz2}} and the New Testament in Tatar. The version of the translation tested was {\tt {\small r37137}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{lrr}
  \toprule
   Corpus                  & Tokens    & Coverage\\
	\midrule
   Tatar New Test.         & 163,603   & 72.04\% \\
   Tatar Wikipedia         & 37,123    & 70.19\% \\
   \midrule
   Bashkir Wikipedia       & 12,267    & 65.99\% \\
   \bottomrule
  \end{tabular}
    \caption{Na\"ive vocabulary coverage over the three corpora.}
 %\vspace{-1.5em}
    \label{table:coverage}
  \end{center}
\end{table}

As shown in Table~\ref{table:coverage}, the coverage is still far too low to be of use as a 
general broad-domain MT system, but we hope that it shows that a good proportion of 
the morphology of both languages is in place.

To get an idea of the kind of performance that could be expected from the system, we 
translated a simple story from Tatar to Bashkir and vice versa. The story may be found 
online,\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/xupaixkar/rasskaz}}
and was used for pedagogical purposes in a recently workshop on MT
for the languages of Russia.

\begin{table}[htbp]
  \begin{center}
  \begin{tabular}{ccrrr}
  \toprule
   Corpus                 & Direction         & Tokens  & Unknown & WER  \\
  \midrule
   \multirow{2}{*}{story} & tt$\rightarrow$ba & 311     & 9  & 8.97\% \\
                          & ba$\rightarrow$tt & 312     & 1  & 7.72\%  \\
  \bottomrule
  \end{tabular}
    \caption{Word error rate and over the small test corpus.}
 %\vspace{-2.5em}
    \label{table:wer}
  \end{center}
\end{table}

Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
distance \citep{levenshtein/1966}. This measure was calculated once all the stems in the 
text had been added to the system, thus presents an upper bound on the current performance
of the transfer lexicon, and the disambiguation and transfer rules. The difference in 
the number of unknown words between translating Tatar$\rightarrow$Bashkir and vice versa
is because certain forms were not found due to lack of corresponding morphophonological rules.

We calculate the WER instead of other MT evaluation metrics such as BLEU as the WER is 
geared towards a particular task, that of measuring postedition effort. The translations 
of the story into Tatar and Bashkir were done in parallel to make them as close as possible,
so using BLEU would give an over-optimistic view of the quality.

\subsection{Error analysis}

The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
\begin{itemise}
  \item Combinations of case and possessive suffixes,
  \item Orthographical representations of phonology,
  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
\end{itemise}

%\cite{TyersAlperen2010}

%challenges:

%% making bidirectional systems
%% when the morphotactics does not line up

\section{Concluding remarks}
\label{sec:conc}

To our knowledge we have presented the first ever MT system between Tatar and Bashkir, and the first ever MT system involving Bashkir. The system is available as free/open-source software under the GNU GPL and the 
whole system may be downloaded from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/nursery/apertium-tt-ba}}

We plan to continue development on the pair; the main work will be expanding the dictionaries with new lists of stems, and providing bilingual correspondences. The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.

\section*{Acknowledgements}

We would like to thank the anonymous reviewers for their helpful comments in improving the paper. This work has been partially funded by Spanish Ministerio de Ciencia e Innovación through project TIN2009-14009-C02-01.

\bibliographystyle{lrec2012}
\bibliography{tt-ba}

\end{document}

### 2013-02-26 paper/paper.tex
remove/rename some of the Bashkir related stuff

@@
 \documentclass[11pt,a4paper]{article}
-%\usepackage{lrec2012}
+\usepackage{mtsxiv}
 %\usepackage{url}
@@
 
-\title{A prototype machine translation system for Tatar and Bashkir based on 
+\title{A prototype machine translation system for Kazakh and Tatar based on 
    free/open-source components}
@@
 This paper presents a prototype bidirectional machine translation system between 
-Tatar and Bashkir, two minority Turkic languages of Russia. While the 
-system has low open-domain coverage, results are presented that suggest
-that high accuracy may be obtained between these two closely-related
-languages, on a par with similar systems.
+Kazakh and Tatar.
  \\
-\Keywords{Tatar, Bashkir, MT, Free software, Open-source}
+\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
 }
@@
 This paper presents a prototype shallow-transfer rule-based machine translation
-system between Tatar and Bashkir, two closely-related minority Turkic 
-languages of Russia.
+system between Kazakh and Tatar.
 

### 2013-02-26 paper/paper.tex
add few sentences about Kazakh

@@
 work in the area of Turkic--Turkic language translation; Section~\ref{sec:sys}\ 
-describes the system and the tools used to construct it; Section~\ref{sec:eval}\ gives a very preliminary evaluation of the system; and finally Section~\ref{sec:conc}\ describes our aims for future work and some concluding remarks.
+describes the system and the tools used to construct it; Section~\ref{sec:eval}\ gives
+a preliminary evaluation of the system; and finally Section~\ref{sec:conc}\ describes
+our aims for future work and some concluding remarks.
 
@@
 
-Tatar is a Turkic language spoken in and around Tatarstan by approximately 6 million people.
-Bashkir (Bashqort) is a Turkic language spoken by about 1.5 million people in and around Bashqortostan.
-Tatarstan and Bashqortostan are both republics within Russia.
-Both languages are co-official with Russian in their respective republics.
-The two languages belong to the same branch of the Kypchak group of Turkic languages.
-As they are very close relatives, they share many innovations, but Bashkir has quite a few phonological innovations beyond those of Tatar (such as rounding harmony and desonorisation of high-sonority suffix-initial consonants; cf.\ \citet{washington10}) and the languages have a number of morphological differences (including different volitional participles).
-The spoken languages share a high level of mutual intelligibility, but many of the inherent similarities are obscured by their fairly different orthographical systems along with the phonological and morphological differences between the languages.
-
-Aside from native speaker intuition, we also consulted the Bashkir Grammar of M. G. Usmanova \citep{usmanova06}.
+Both languages belong to the Kypchak (or Northwestern) group of Turkic languages.
+Kazakh is primarily spoken in Kazakhstan, where it is the national language. Large
+groups of native speakers also exist in neigbouring Central-Eurasian republics,
+China and Mongolia. Total number of speakers is at least 10 million people.
 

### 2013-02-27 paper/paper.tex
add few words about Tatar
some more

@@
 
-\title{A prototype machine translation system for Kazakh and Tatar based on 
-   free/open-source components}
+\title{A free/open-source Kazakh-Tatar machine translation system}
 
@@
 \abstract{
-This paper presents a prototype bidirectional machine translation system between 
+This paper presents a bidirectional machine translation system between
 Kazakh and Tatar.
@@
 
-The paper will be laid out as follows: Section~\ref{sec:lang}\ gives a brief description
-of the two languages; Section~\ref{sec:prev}\ gives a short review of some previous
-work in the area of Turkic--Turkic language translation; Section~\ref{sec:sys}\ 
-describes the system and the tools used to construct it; Section~\ref{sec:eval}\ gives
-a preliminary evaluation of the system; and finally Section~\ref{sec:conc}\ describes
-our aims for future work and some concluding remarks.
+The paper will be laid out as follows: Section~\ref{sec:lang}\ gives a brief
+description of the two languages; Section~\ref{sec:prev}\ gives a short review
+of some previous work in the area of Turkic--Turkic language translation;
+Section~\ref{sec:sys}\ describes the system and the tools used to construct it;
+Section~\ref{sec:eval}\ gives a preliminary evaluation of the system; and
+finally Section~\ref{sec:conc}\ describes our aims for future work and some
+concluding remarks.
 
@@
 
-Both languages belong to the Kypchak (or Northwestern) group of Turkic languages.
-Kazakh is primarily spoken in Kazakhstan, where it is the national language. Large
-groups of native speakers also exist in neigbouring Central-Eurasian republics,
-China and Mongolia. Total number of speakers is at least 10 million people.
+Both languages belong to the Kypchak (or Northwestern) group of Turkic
+languages. Kazakh is primarily spoken in Kazakhstan, where it is the national
+language. Large groups of native speakers also exist in China, neigbouring
+Central-Eurasian republics and Mongolia. Total number of speakers is at least
+10 million people. Tatar is a Turkic language spoken in and around Tatarstan by
+approximately 6 million people. It is co-official with Russian in Tatarstan --
+a republic within Russia.
 
@@
 Turkic languages exist, although to our knowledge none are publically available.
+For systems between Turkish and other Turkic languages, there have been,
+for example, systems reported for Turkish-Crimean Tatar \citep{altintas01},
+Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and
+Turkish-Turkmen \citep{tantug07}.
 
-For systems between Turkish and other Turkic languages, there have been, 
-for example, systems reported for Turkish-Crimean Tatar \citep{altintas01}, Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and Turkish-Turkmen \citep{tantug07}.
+A prototype Tatar-Bashkir machine translation system was built by ourselves.
+Due to the closeness of the languages, it proved to provide high accuracy
+translations, but being a prototype system by design, had relatively low
+coverage.
 

### 2013-02-27 paper/paper.tex
some more
humble myself and put my name at the end

@@
 
-\name{Ilnar Salimzyanov, Jonathan North Washington, and Francis M. Tyers}
+\name{Jonathan North Washington, Francis M. Tyers and Ilnar Salimzyanov}
 
-\address{ Kazan Federal University, Indiana University, Universitat d'Alacant \\
-          Kazan, Bloomington, Alacant\\
-          Russia, IN 47405 (USA), E-03071 Spain\\
-          \email{ilnar.salimzyan@gmail.com}, \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es} \\
+\address{ Indiana University, Universitat d'Alacant, Kazan Federal University \\
+          Bloomington, Alacant, Kazan\\
+          IN 47405 (USA), E-03071 Spain, Russia\\
+          \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es},  \email{ilnar.salimzyan@gmail.com} \\
 }
@@
 coverage.
+
+%maybe this should be moved to 'Future plans'
+Inside the Apertium project, work on several Turkic languages started to be
+carried out, with the ultimate goal of creating independent but compatible
+finite-state transducers for each language.
+%% The idea is that we started Kazakh-Tatar pair to be able to take into account
+%% characteristics of two more distant Turkic languages, avoiding narrowing
+%% ourselves to mutually intellegible Tatar and Bashkir, but generate
+%% transducers which are still compatible and very coherent. Bashkir part of the
+%% pair will be updated to the latest developments of Tatar transducer, which
+%% took place while working on Kazakh-Tatar.
 

### 2013-04-06 paper/paper.tex
a little poking

@@
 \usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=blue]{hyperref}
-\usepackage{times}
+%\usepackage{times}
 \usepackage{multirow}
@@
 
-Both languages belong to the Kypchak (or Northwestern) group of Turkic
+Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
 languages. Kazakh is primarily spoken in Kazakhstan, where it is the national
 language. Large groups of native speakers also exist in China, neigbouring
-Central-Eurasian republics and Mongolia. Total number of speakers is at least
+Central-Eurasian republics, and Mongolia. The total number of speakers is at least
 10 million people. Tatar is a Turkic language spoken in and around Tatarstan by
 approximately 6 million people. It is co-official with Russian in Tatarstan --
-a republic within Russia.
+a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.
 
@@
 Turkic languages exist, although to our knowledge none are publically available.
-For systems between Turkish and other Turkic languages, there have been,
-for example, systems reported for Turkish-Crimean Tatar \citep{altintas01},
+%FIXME: tr-az, tr-ky
+For systems between Turkish and other Turkic languages, there have been systems reported for Turkish-Crimean Tatar \citep{altintas01},
 Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and
-Turkish-Turkmen \citep{tantug07}.
+Turkish-Turkmen \citep{tantug07}, though none of these have been released to a public audience.
 
-A prototype Tatar-Bashkir machine translation system was built by ourselves.
+A prototype Tatar-Bashkir machine translation system was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}.
 Due to the closeness of the languages, it proved to provide high accuracy
@@
 
-\bibliographystyle{lrec2012}
-\bibliography{tt-ba}
+%\bibliographystyle{lrec2012}
+%\bibliography{tt-ba}
+\bibliographystyle{apa}
+\bibliography{paper}
 

### 2013-04-06 paper/paper.tex
more poking on previous work section

@@
 
-Several previous works on making machine translation systems between
-Turkic languages exist, although to our knowledge none are publically available.
-%FIXME: tr-az, tr-ky
-For systems between Turkish and other Turkic languages, there have been systems reported for Turkish-Crimean Tatar \citep{altintas01},
+Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish-Kyrgyz, Azeri-Turkish), but the Kazakh-Tatar system described by the present study is the closest to production-ready of them.
+Among these systems is a prototype Tatar-Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
+
+Besides these systems, several previous works on making machine translation systems between Turkic languages exist, although to our knowledge none are publically available.
+Some MT systems have been reported that translate between Turkish and other Turkic languages, including Turkish-Crimean Tatar \citep{altintas01},
 Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and
@@
 
-A prototype Tatar-Bashkir machine translation system was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}.
-Due to the closeness of the languages, it proved to provide high accuracy
-translations, but being a prototype system by design, had relatively low
-coverage.
+%FIXME: mention tr-az, tr-ky?
 
 %maybe this should be moved to 'Future plans'
-Inside the Apertium project, work on several Turkic languages started to be
-carried out, with the ultimate goal of creating independent but compatible
+Within the Apertium project, work on several Turkic-language MT systems has been started, with the ultimate goal of creating independent but compatible
 finite-state transducers for each language.

### 2013-04-06 paper/paper.tex
less ERROR

@@
 
-\name{Jonathan North Washington, Francis M. Tyers and Ilnar Salimzyanov}
+%FIXME
+%\name{Jonathan North Washington, Francis M. Tyers and Ilnar Salimzyanov}
 
-\address{ Indiana University, Universitat d'Alacant, Kazan Federal University \\
-          Bloomington, Alacant, Kazan\\
-          IN 47405 (USA), E-03071 Spain, Russia\\
-          \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es},  \email{ilnar.salimzyan@gmail.com} \\
-}
+%\address{ Indiana University, Universitat d'Alacant, Kazan Federal University \\
+%          Bloomington, Alacant, Kazan\\
+%          IN 47405 (USA), E-03071 Spain, Russia\\
+%          \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es},  \email{ilnar.salimzyan@gmail.com} \\
+%}
 
@@
  \\
-\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
+%FIXME
+%\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
 }
@@
 \spacing{1}
-\maketitleabstract
+%FIXME
+%\maketitleabstract
 

### 2013-04-06 paper/paper.tex
MUCH BETTAR

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[11pt]{article}
+
 \usepackage{mtsxiv}
@@
 \usepackage{multirow}
+\setlength\titlebox{6.5cm}    % Expanding the titlebox
+
 \usepackage{polyglossia}
@@
 
-\title{A free/open-source Kazakh-Tatar machine translation system}
-
 %FIXME
@@
 
+%\abstract{
+%This paper presents a bidirectional machine translation system between
+%Kazakh and Tatar.
+% \\
+%%FIXME
+%%\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
+%}
+
+\newcommand{\confname}{Machine Translation Summit XIV}
+\newcommand{\website}{\protect\url{http://www.mtsummit2013.info/}}
+\newcommand{\contactname}{research track co-chair Mikel L.\ Forcada}
+\newcommand{\contactemail}{mlf@dlsi.ua.es} 
+\newcommand{\conffilename}{mtsxiv}
+\newcommand{\downloadsite}{\protect\url{http://www.mtsummit2013.info/}}
+\newcommand{\paperlength}{$8$ (eight)}
+\newcommand{\shortpaperlength}{$4$ (four)}
+
+\title{A free/open-source Kazakh-Tatar machine translation system}
+
+\author{Ilnar Salimzyanov\\
+  Affiliation / Address line 1\\
+  Affiliation / Address line 2\\
+  Affiliation / Address line 3\\
+  {\tt email@domain}  \And
+  Jonathan North Washington\\
+  Affiliation / Address line 1\\
+  Affiliation / Address line 2\\
+  Affiliation / Address line 3\\
+  {\tt email@domain}  \And
+  Francis Morton Tyers\\
+  Affiliation / Address line 1\\
+  Affiliation / Address line 2\\
+  Affiliation / Address line 3\\
+  {\tt email@domain}}
+
 \date{\today}
-
-\abstract{
-This paper presents a bidirectional machine translation system between
-Kazakh and Tatar.
- \\
-%FIXME
-%\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
-}
 
@@
 \spacing{1}
+\maketitle
+\begin{abstract}
+This paper presents a bidirectional machine translation system between Kazakh and Tatar.
+\end{abstract}
 %FIXME
@@
 %\bibliography{tt-ba}
-\bibliographystyle{apa}
+\bibliographystyle{mtsxiv}
 \bibliography{paper}

### 2013-04-06 paper/paper.tex
NO MOAR ERRORS
cleaner

@@
 \newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}
+\setmonofont[Scale=0.7]{DejaVu Sans Mono}
 
@@
 
-%FIXME
-%\name{Jonathan North Washington, Francis M. Tyers and Ilnar Salimzyanov}
-
-%\address{ Indiana University, Universitat d'Alacant, Kazan Federal University \\
-%          Bloomington, Alacant, Kazan\\
-%          IN 47405 (USA), E-03071 Spain, Russia\\
-%          \email{jonwashi@indiana.edu}, \email{ftyers@dlsi.ua.es},  \email{ilnar.salimzyan@gmail.com} \\
-%}
-
 \newcommand{\tag}[1]{{\small{\texttt{#1}}}}
 
-%\abstract{
-%This paper presents a bidirectional machine translation system between
-%Kazakh and Tatar.
-% \\
-%%FIXME
 %%\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
-%}
 
@@
   Affiliation / Address line 3\\
-  {\tt email@domain}  \And
+  \email{email@domain}  \And
   Jonathan North Washington\\
-  Affiliation / Address line 1\\
-  Affiliation / Address line 2\\
-  Affiliation / Address line 3\\
-  {\tt email@domain}  \And
+  Departments of Linguistics\\
+  and Central Eurasian Studies\\
+  Indiana University\\
+  Bloomington, Indiana\\
+  \email{jonwashi@indiana.edu}  \And
   Francis Morton Tyers\\
@@
   Affiliation / Address line 3\\
-  {\tt email@domain}}
+  \email{email@domain}}
 
@@
 
-\section{Languages}
-\label{sec:lang}
-
-Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
-languages. Kazakh is primarily spoken in Kazakhstan, where it is the national
-language. Large groups of native speakers also exist in China, neigbouring
-Central-Eurasian republics, and Mongolia. The total number of speakers is at least
-10 million people. Tatar is a Turkic language spoken in and around Tatarstan by
-approximately 6 million people. It is co-official with Russian in Tatarstan --
-a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.
-
 \section{Previous work}
@@
 %% took place while working on Kazakh-Tatar.
+
+\section{Languages}
+\label{sec:lang}
+
+Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
+languages. Kazakh is primarily spoken in Kazakhstan, where it is the national
+language. Large groups of native speakers also exist in China, neigbouring
+Central-Eurasian republics, and Mongolia. The total number of speakers is at least
+10 million people. Tatar is a Turkic language spoken in and around Tatarstan by
+approximately 6 million people. It is co-official with Russian in Tatarstan --
+a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.
+
+%FIXME: Contrastive grammar
 

### 2013-04-06 paper/paper.tex
beginnings of comparative grammar

@@
 Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
-languages. Kazakh is primarily spoken in Kazakhstan, where it is the national
-language. Large groups of native speakers also exist in China, neigbouring
-Central-Eurasian republics, and Mongolia. The total number of speakers is at least
-10 million people. Tatar is a Turkic language spoken in and around Tatarstan by
+languages.  They share some level of mutual intelligibility to naïve native speakers, though this is somewhat limited, especially regarding the written languages.
+
+Kazakh is primarily spoken in Kazakhstan, where it is the national
+language, sharing official status with Russian as the FIXME language.
+Large groups of native speakers also exist in China, neigbouring
+Central-Eurasian republics, and Mongolia. The total number of speakers is at least 10 million people.
+
+Tatar is spoken in and around Tatarstan by
 approximately 6 million people. It is co-official with Russian in Tatarstan --
@@
 %FIXME: Contrastive grammar
+\subsection{Phonological differences}
+%a generalised summary, 2 or 3 small specific examples
+As closely related languages, Kazakh and Tatar share many phonological processes, including front-back vowel harmony systems, consonant voicing assimilation, and even a typologically rare consonantal nasal harmony system.
+However, the differing details of these processes and the existence of processes unique to each language render Kazakh and Tatar fairly different.
+For example, Kazakh has a ubiquitous system of desonorisation of the initial sonorants found in many common morphemes.
+Furthermore, Tatar has nasal assimilation of the initial /l/ of the plural-suffix.
+
+\subsection{Orthographic differences}
+a generalised summary, 1 or 2 small specific examples
+\subsection{Morphotactic differences}
+2 or 3 specific examples
+\subsection{Lexical and morphological differences}
+2 or 3 specific examples
+\subsection{Syntactic differences}
+2 or 3 specific examples
+> ұна- vs.\ ярат- (does this count as syntactic?)
 

### 2013-04-06 paper/paper.tex
-- to ---

@@
 Tatar is spoken in and around Tatarstan by
-approximately 6 million people. It is co-official with Russian in Tatarstan --
+approximately 6 million people. It is co-official with Russian in Tatarstan ---
 a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.
@@
 It also supports other finite state transducer formalisms such as \textbf{sfst}.
-This toolkit has been chosen as it -- or the equivalent XFST -- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006}, and is available under a free/open-source licence.
+This toolkit has been chosen as it --- or the equivalent XFST --- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006}, and is available under a free/open-source licence.
 

### 2013-04-06 paper/paper.tex
add affil.
affiliations

@@
   Indiana University\\
-  Bloomington, Indiana\\
+  Bloomington, Indiana 47405 USA\\
   \email{jonwashi@indiana.edu}  \And
   Francis Morton Tyers\\
-  Affiliation / Address line 1\\
-  Affiliation / Address line 2\\
-  Affiliation / Address line 3\\
-  \email{email@domain}}
+  Departament de Llenguatges\\
+  i Sistemes Informàtics\\
+  Universitat d'Alacant\\
+  E-03877 Alacant\\
+  \email{ftyers@dlsi.ua.es}}
 

### 2013-04-07 paper/paper.tex
some orthographic stuff

@@
 approximately 6 million people. It is co-official with Russian in Tatarstan ---
-a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian.
+a republic within Russia.  A majority of native speakers of both languages are bilingual in Russian. %, and the standard varieties our system FIXMEs are written in Cyrillic.
 
@@
 \subsection{Orthographic differences}
-a generalised summary, 1 or 2 small specific examples
+%a generalised summary, 1 or 2 small specific examples
+The standard varieties of Kazakh and Tatar our system deals with are both written in Cyrillic, though their implementations of Cyrillic differ in many ways.
+
+While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent vowels (and employing ``hard'' and ``soft signs'' when these fail) to differentiate the two, and Kazakh has two series of obstruents (e.g., ‹к› and ‹қ›).
+
+Kazakh does not orthographically distinguish high unrounded vowels (/ɘ/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /ɘj/ and /əj/ are both written ‹и›, while /ɘw/ and /əw/ are both written ‹у›.
+The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels, so this convention presents accute challenges to producing an accurate morphological transducer.
+
+Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›.}
+
 \subsection{Morphotactic differences}

### 2013-04-07 paper/paper.tex
a little polishing of orthography section

@@
 
-Kazakh does not orthographically distinguish high unrounded vowels (/ɘ/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /ɘj/ and /əj/ are both written ‹и›, while /ɘw/ and /əw/ are both written ‹у›.
-The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels, so this convention presents accute challenges to producing an accurate morphological transducer.
+Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.
+The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels.
+%, so this convention presents accute challenges to producing an accurate morphological transducer.
+Additionally, Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,
 
-Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›.}
+All of these orthographical conventions present accute challenges to designing accurate morphological transducers for the languages.
 

### 2013-04-07 paper/paper.tex
a syntactic difference

@@
 \subsection{Syntactic differences}
-2 or 3 specific examples
-> ұна- vs.\ ярат- (does this count as syntactic?)
+%2 or 3 specific examples
+There aer a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.
+
+%FIXME: Ilnar, please check the Tatar
+An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.
+In Kazakh, the verb ‹ұна› is used, e.g.
+‹оқу маған ұнайды› ``I like studying'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``study'' in English is the grammatical subject in Kazakh.
+Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g.
+‹мин укуны яратам› ``I like studying'', where the first person pronoun is the grammatical subject and ``studying'' is the grammatical direct object.
+%(does this count as syntactic?)
 

### 2013-04-07 paper/paper.tex
stupid typo

@@
 %2 or 3 specific examples
-There aer a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.
+There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.
 

### 2013-04-13 paper/paper.tex
need Ilnar's help
adding affiliation

@@
 \author{Ilnar Salimzyanov\\
-  Affiliation / Address line 1\\
-  Affiliation / Address line 2\\
-  Affiliation / Address line 3\\
-  \email{email@domain}  \And
+  Kazan Federal University\\
+  Kazan, Republic of Tatarstan\\
+  Russian Federation\\
+  \email{ilnar.salimzyan@gmail.com}  \And
   Jonathan North Washington\\
@@
 2 or 3 specific examples
+%FIXME: Ilnar, I can't think of anything for this...
+
 \subsection{Lexical and morphological differences}
 2 or 3 specific examples
+> present progressive
+%FIXME: Ilnar, could you come up with some good examples of this?
+> ... ?
+
 \subsection{Syntactic differences}

### 2013-04-16 paper/paper.tex
some examples for contrastive analysis of the languages, need elaboraton
some restructuring of comparitive stuff based on Ilnar's input
new bib entry, but now broken compile :(

@@
 
-\subsection{Morphotactic differences}
+\subsection{Morphological differences}
 2 or 3 specific examples
-%FIXME: Ilnar, I can't think of anything for this...
 
-\subsection{Lexical and morphological differences}
-2 or 3 specific examples
-> present progressive
-%FIXME: Ilnar, could you come up with some good examples of this?
-> ... ?
+> Possessive endings system (yeah!)
+> Kazakh -EtIn
+> Past tense system
+
+%\subsection{Morphotactic differences}
+%2 or 3 specific examples
+%
+%%FIXME: Ilnar, I can't think of anything for this...
+%% The greetings in Tatar kept a probably older variant of /personal copula ending - question
+%% partice/ sequence (Исәнмесез vs Сәлеметсіз бе?), but this doesn't seem to be a good example,
+%% as it is very "special case".
+%% Although, the possessive endings system is somewhat different, maybe there is something.
+
+%\subsection{Lexical and morphological differences}
+%2 or 3 specific examples
+%> present progressive
+%> ... ?
+%
+%%FIXME: Ilnar, could you come up with some good examples of this?
+%% * no instrumental case in Tatar
+%% * only one formal 2p pronoun in Tatar (i.e. Сез) as opposed to Kazakh "Сіз" and "Сіздер"
+%% * барыпты - барган (икән), сөйлескелі келдім - сөйләшергә килдем, баратын - бара торган
 
 \subsection{Syntactic differences}
+
 %2 or 3 specific examples
+% * Different case governance of postpositions (grep for WARNINGs in bidix)
+% * музейлік экспозиция > музей экспозициясе; экспозициялық зал > экспозиция залы
+% ** This works, there is a rule for that: if an /adjective + noun/ is translated with /n.attr + noun/, <px3sp> affix is added in Tatar 
+% * Мен оқуды жақсы көремін > Мин укырга яратам.
+% Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
 There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.
@@
 %FIXME: Ilnar, please check the Tatar
+% There is "ошый" in Tatar, which works the same way as Kazakh verb.
+% But I found the following in the bidix:
+
+% <e>       <p><l>иемден<s n="v"/><s n="tv"/></l>   <r>ия<b/>бул<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->
+% <e>       <p><l>сақтандыр<s n="v"/><s n="iv"/></l><r>кисәт<s n="v"/><s n="tv"/></r></p></e> <!--"" WARNING: IV<>TV-->
+% <e>       <p><l>тұтын<s n="v"/><s n="tv"/></l>    <r>тотын<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->
+% <e>       <p><l>тұтын<s n="v"/><s n="tv"/></l>    <r>тотын<s n="v"/><s n="iv"/></r></p></e> <!--"" WARNING transitivity-->
+
+% There are few other WARNINGs as well.
+
+%FIXME: use a noun in examples, not infinitive/gerund
 An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.
@@
 %(does this count as syntactic?)
+
+> Tatar infinitive (Kazakh ger+nom/acc/dat, Kazakh -GAlI vadv)
 
@@
 It also supports other finite state transducer formalisms such as \textbf{sfst}.
-This toolkit has been chosen as it --- or the equivalent XFST --- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006}, and is available under a free/open-source licence.
+This toolkit has been chosen as it --- or the equivalent XFST --- has been widely used for other Turkic languages \citep{coltekin2010,altintas2001,tantug2006,washingtonipasovtyers12,tyerswashingtonsalimzyanbattalov12}, and is available under a free/open-source licence.
 

### 2013-04-16 paper/paper.tex
work-around for MT Summit's broken bst file
Morphological Transducers (section 4.2): check

@@
 Kazakh is primarily spoken in Kazakhstan, where it is the national
-language, sharing official status with Russian as the FIXME language.
+language, sharing official status with Russian as an official language.
 Large groups of native speakers also exist in China, neigbouring
@@
 Use of lexc allows for straightforward definition of different word classes and subclasses.
-For example, Tatar (but not Bashkir) has two classes of verbs: one which take a harmonised high vowel in the infinitive (the default), and one which take a harmonised low vowel in the infinitive.
+For example, Tatar (but not Kazakh) has two classes of verbs: one which takes a harmonised high vowel in the infinitive (the default), and one which take a harmonised low vowel in the infinitive.
 This was implemented in lexc with two similar continuation lexica for verbs: one pointing at a lexicon with an A-initial infinitive ending, and another pointing at a lexicon with an I-initial infinitive ending.
@@
 \subsection{Bilingual lexicon}
+%FIXME: Ilnar, we need some good examples here
 \begin{figure*}[htbp]
@@
 \bibliographystyle{mtsxiv}
+%\bibliographystyle{apa}
 \bibliography{paper}

### 2013-04-17 paper/paper.tex
today's work

@@
 \usepackage{booktabs}
+
+%\usepackage{subfigure}
+%\usepackage{subfloat}
+\usepackage{subfig}
 
@@
 Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
-languages.  They share some level of mutual intelligibility to naïve native speakers, though this is somewhat limited, especially regarding the written languages.
+languages.
+The spoken and written languages share some level of mutual intelligibility to naïve native speakers, though this is somewhat limited, and is obscured by different orthographical conventions and some opaque correspondences.
 
@@
 \subsection{Morphological differences}
-2 or 3 specific examples
+There are a number of examples where the morphologies of Kazakh and Tatar are rather different, including morphemes in one language that don't exist in the other, entirely different uses of the same morpheme combinations, and morphotactic differences (i.e., allowable ordering and placement of morphemes).
 
-> Possessive endings system (yeah!)
-> Kazakh -EtIn
+An example of a morpheme that doesn't exist in one of the two languages is Kazakh -\texttt{\{E\}т\{I\}н}, which is used to form non-past verbal adjectives and verbal nouns.
+The semantically equivalent structure in Tatar is -\texttt{\{E\} торган}, which historically corresponds to the source of the Kazakh morpheme; however, the use of -\texttt{\{E\} тұрған} in modern Kazakh is different from that of -\texttt{\{E\}т\{I\}н}.
+
+%Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.
+Kazakh has a distinct pronoun for all combinations of [±plural, ±formal] (сен, сендер, сіз, сіздер), whereas Tatar collapse all pronouns except the [-plural, -formal] син into one pronoun (сез).
+This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
+Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.
+Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
+\begin{table}[htbp]
+%\begin{subtables}
+	%\begin{subtable}[l]
+	%\subtable[Kazakh 2nd person pronouns]{
+	%\begin{subtable}
+	\subfloat[Kazakh 2nd person pronouns]{
+		\begin{tabular}{lll}
+			\toprule
+			 & -pl & +pl \\\midrule
+			-formal & сен & сендер \\
+			+formal & сіз & сіздер \\\bottomrule
+		\end{tabular}
+	%	\caption{Kazakh 2nd person pronouns}
+	%\end{subtable}
+	%\end{subtable}
+	%\begin{subtable}
+	%\begin{subtable}
+	%}\subtable[Tatar 2nd person pronouns]{
+	}
+	\subfloat[Tatar 2nd person pronouns]{
+		\begin{tabular}{lll}
+			\toprule
+			 & -pl & +pl \\\midrule
+			-formal & син & сез \\
+			+formal & сез & сез \\\bottomrule
+		\end{tabular}%}
+	%	\caption{Kazakh 2nd person pronouns}
+	%\end{subtable}
+	%\end{subtable}
+	}
+	\caption{The 2nd person pronoun systems of Kazakh and Tatar}
+	\label{tab:2pprons}
+\end{table}
+%\end{subtables}
+
 > Past tense system

### 2013-04-17 paper/paper.tex
a little more space between the tables
some clarification about morphemes

@@
 Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.
-Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
+Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}
 \begin{table}[htbp]
@@
 	%}\subtable[Tatar 2nd person pronouns]{
-	}
+	}\vspace{1.5ex}
 	\subfloat[Tatar 2nd person pronouns]{
@@
 %\end{subtables}
+This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.
+However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.
+Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language.
 

### 2013-04-17 paper/paper.tex
some examples from the bidix; add the number of stems

@@
 %FIXME: Ilnar, we need some good examples here
+% ақпараттық құрал - мәгълүмат чарасы
+% құрал - корал in other contexts
+
+%<e>       <p><l>жай<s n="n"/></l>                 <r>җәя<s n="n"/></r></p></e> <!--"arch"-->
+%<e>       <p><l>жай<s n="n"/></l>                 <r>урын<s n="n"/></r></p></e> <!--"place"-->
+%<e>       <p><l>жай<s n="n"/></l>                 <r>хәл<s n="n"/></r></p></e> <!--"condition"-->
+%<e>       <p><l>жай<s n="n"/></l>                 <r>яшен<s n="n"/></r></p></e> <!--"lightning"-->
+
 \begin{figure*}[htbp]
 \begin{texttt}
-    <e><p><l>\textbf{юнәлеш}<s n="n"/></l><r>\textbf{йүнәлеш}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{танау}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
     <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
@@
 \end{texttt}
-\caption{Example entries from the bilingual transfer lexicon. Tatar is on the left, and Bashkir on the right}
+\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
 \label{fig:bidix}
@@
 
-The bilingual lexicon currently contains 2,834 stem to stem correspondences and was build by hand by a bilingual 
-speaker of Tatar and Bashkir, translating a frequency list of the Russian National Corpus\footnote{\url{http://ruscorpora.ru/en/}} into both languages in a spreadsheet.
-This spreadsheet was then converted into the Apertium
-XML dictionary format.
+% FIXME: the following sh... stuff needs to be rephrased
+The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand.
+Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - "автомобиль" etc).
 

### 2013-04-17 paper/paper.tex
past tense semantic differences

@@
 
+The past tense systems of Kazakh and Tatar also do not correspond.
+At a basic level, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may equate more to sureness of knowledge.} (where [-eyewitness] is used in cases of newly discovered information) and [±recent] in the past tense, whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information ---all with no [±recent] distinction).
+The result of this is a many-to-many correspondence.
+For example, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн}) while Kazakh has a recent non-eyewitness past (\texttt{-IптI}) and a distant non-eyewitness past (\texttt{-GAн екен-}).
+On the other hand, these two non-eyewitness past forms are used for newly acquired information in Kazakh, whereas in Tatar non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past tenses are distinguished.
 > Past tense system

### 2013-04-17 paper/paper.tex
cleaned up bit about past tense systems

--- 2013-04-17 paper/paper.tex
@@
 
-> Past tense system
+The past tense systems of Kazakh and Tatar have a many-to-many correspondence.
+At a basic level, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent] in the past tense, whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.
+As an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн}) while Kazakh has a recent non-eyewitness past (\texttt{-IптI}) and a distant non-eyewitness past (\texttt{-GAн екен-}).
+On the other hand, these two non-eyewitness past forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
+
+Morphotactics of -GAн екен/икән
 

### 2013-04-17 paper/paper.tex
a bunch of copy-editing

--- 2013-04-17 paper/paper.tex
@@
 This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
-Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.
-Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}
+Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a four-way distinction in Kazakh's 2nd person system (both pronouns and agreement suffixes), where Tatar only has a two-way distinction.
+Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}.
 \begin{table}[htbp]
@@
 %\end{subtables}
+
 This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.
 However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.
-Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language.
+Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.
 
-> Past tense system
+The past tense systems of Kazakh and Tatar have a many-to-many correspondence.
+At a basic level, in the past tense, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.
+As an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a distant non-eyewitness past (\texttt{-GAн екен-}).
+On the other hand, these two non-eyewitness past forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
+
+Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.
+Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').
+
+Another morphotactic difference between Kazakh and Tatar is found with the negative forms of the cognate \texttt{-GAн-} past tenses.
+In Kazakh, the negative form of the non-recent reliable-information past tense is \texttt{-GAн емес-}, whereas in Tatar, the negative form of the non-eyewitness past tense is \texttt{-мAGAн-}.
+
+%Morphotactics of -GAн екен/икән
 

### 2013-04-18 paper/paper.tex
maybe something like this for the tables?  spectie, selimcan, please have a look

@@
 On the other hand, these two non-eyewitness past forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
+
+\begin{table}[htbp]
+	\centering
+	\subfloat[Kazakh past tense morphology]{
+		\begin{tabular}{lll}
+			\toprule
+			 & [+recent] & [-recent] \\\midrule
+			{}[+reliable] & -DI- & -GAн- \\
+			{}[-reliable] & -Iп- & -GAн екен- \\ \bottomrule
+		\end{tabular}
+%		\begin{tabular}{lll}
+%			\toprule
+%			 & recent & non-recent \\\midrule
+%			eyewitness/reliable & -DI- & -GAн- \\
+%			non-eyewitness & \multirow{2}*{-Iп-} & \multirow{2}*{-GAн екен-} \\
+%			{} / newly acquired information & & \\ \bottomrule
+%		\end{tabular}
+
+	}\\
+	\subfloat[Tatar past tense morphology]{
+		\begin{tabular}{lll}
+			\toprule
+			 & [-newlyAcq'd] & [+newlyAcq'd] \\\midrule
+			{}[+reliable] & -DI- & \multirow{2}*{-GAн- икән}\\
+			{}[-reliable] & -GAн- & \\\bottomrule
+		\end{tabular}
+%		\begin{tabular}{lll}
+%			\toprule
+%			 & past \\\midrule
+%			eyewitness & -DI- \\
+%			non-eyewitness/hearsay & -GAн- \\
+%			newly acquired information & -GAн- икән \\\bottomrule
+%		\end{tabular}
+	}
+	\caption{A comparison of the basic past-tense morphology of Kazakh and Tatar}
+	\label{tab:pasttense}
+\end{table}
 

### 2013-04-18 paper/paper.tex
the last of tonight's work (syntactic differences, new test sentence, etc.)

@@
 \newfontfamily\qipa[IPA,Scale=MatchLowercase]{FreeSerif}
+\newfontfamily\qipb[IPA,Scale=MatchLowercase]{Junicode}
 \setmonofont[Scale=0.7]{DejaVu Sans Mono}
+
+%\fontspec[FakeBold=2.5]{Times New Roman}
+%\DeclareFontShape{EU1}{TimesNewRoman(0)}{m}{sc}{<->ssub * TimesNewRoman(1)/m/sc}{}
 
@@
 \usepackage{booktabs}
+
+\usepackage{gb4e}  % for numbered examples
+\usepackage{expex} % another possibility
 
@@
 \newcommand{\tag}[1]{{\small{\texttt{#1}}}}
+\newcommand{\gmk}[1]{{\qipb\scshape #1}}
+\newcommand{\eng}[1]{`#1'}
 
@@
 language, sharing official status with Russian as an official language.
-Large groups of native speakers also exist in China, neigbouring
+Large groups of native speakers also exist in China, neighbouring
 Central-Eurasian republics, and Mongolia. The total number of speakers is at least 10 million people.
@@
 The past tense systems of Kazakh and Tatar have a many-to-many correspondence.
-At a basic level, in the past tense, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.
+As shown in table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.
 As an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a distant non-eyewitness past (\texttt{-GAн екен-}).
@@
 An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.
-In Kazakh, the verb ‹ұна› is used, e.g.
-‹оқу маған ұнайды› ``I like studying'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``study'' in English is the grammatical subject in Kazakh.
+In Kazakh, the verb ұна is used, e.g.
+бауырсақ маған ұнайды ``I like bawyrsaq'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``bawyrsaq'' in English is the grammatical subject in Kazakh.
 Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g.
-‹мин укуны яратам› ``I like studying'', where the first person pronoun is the grammatical subject and ``studying'' is the grammatical direct object.
+мин бавырсак яратам ``I like bawyrsaq'', where the first person pronoun is the grammatical subject and ``bawyrsaq'' is the grammatical direct object (with no accusative suffix since it is indefinite).
 %(does this count as syntactic?)
+
+In Kazakh, a gerund (i.e., verbal noun) with case marking and sometimes person agreement in the form of possessive suffixes is used to make a verb phrase an argument to certain other main phrases.
+In Tatar, many of these phrases use an invariant infinitival form.
+Some examples are shown below:
+
+%\begin{exe}
+%	\setstretch{0.8}
+%	\ex
+%	\begin{xlist}
+%		\ex Мен үйге баруым керек
+%		\gll мен үй-GA бар-у-Iм керек \\
+%		I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need \\
+%		\glt \eng{I need to go home}
+%
+%		\ex Мин өйгә барырга кирәк
+%		\gll мин өй-GA бар-IргA кирәк \\
+%		I home-\gmk{Dat} go-\gmk{Inf} need \\
+%		\glt \eng{I need to go home}
+%	\end{xlist}
+%\end{exe}
+
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\a
+\begingl
+\gla Мен үйге баруым керек. //
+\glb мен үй-GA бар-у-Iм керек //
+\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
+\glft \eng{I need to go home.} //
+\endgl
+\a
+\begingl
+\gla Мин өйгә барырга кирәк. //
+\glb мин өй-GA бар-IргA кирәк //
+\glb I home-\gmk{Dat} go-\gmk{Inf} need //
+\glft \eng{I need to go home.} //
+\endgl
+\xe
 
@@
 \toprule
-{\bf (Tatar) Input} & Һава бүген бик әйбәт, җылы гына. \\ 
+{\bf (Kazakh) Input} & Ол енді ол дыбысты анығырақ ести бастады \\
+%{\bf (Tatar) Input} & Һава бүген бик әйбәт, җылы гына. \\ 
 \midrule
-{\bf Mor. analysis} & \^{}Һава/һава\tag{<n>}\tag{<attr>}/һава\tag{<n>}\tag{<nom>\$} \^{}бүген/бүген\tag{<adv>\$}  \\
- ~             & \^{}бик/бик\tag{<adv>}/бик\tag{<n>}\tag{<attr>}/бик\tag{<n>}\tag{<nom>\$} \\ 
- ~             & \^{}әйбәт/әйбәт\tag{<adj>}/әйбәт\tag{<adj>}\tag{<subst>}\tag{<nom>\$}\^{},/,\tag{<cm>\$} \\ 
- ~             & \^{}җылы/җылы\tag{<n>}\tag{<attr>}/җылы\tag{<n>}\tag{<nom>}/җылы\tag{<adj>}/җылы\tag{<adj>}\tag{<subst>}\tag{<nom>\$} \\
- ~             & \^{}гына/гына\tag{<postadv>\$}\^{}./.\tag{<sent>\$} \\
+%{\bf Mor. analysis} & \^{}Һава/һава\tag{<n>}\tag{<attr>}/һава\tag{<n>}\tag{<nom>\$} \^{}бүген/бүген\tag{<adv>\$}  \\
+% ~             & \^{}бик/бик\tag{<adv>}/бик\tag{<n>}\tag{<attr>}/бик\tag{<n>}\tag{<nom>\$} \\ 
+% ~             & \^{}әйбәт/әйбәт\tag{<adj>}/әйбәт\tag{<adj>}\tag{<subst>}\tag{<nom>\$}\^{},/,\tag{<cm>\$} \\ 
+% ~             & \^{}җылы/җылы\tag{<n>}\tag{<attr>}/җылы\tag{<n>}\tag{<nom>}/җылы\tag{<adj>}/җылы\tag{<adj>}\tag{<subst>}\tag{<nom>\$} \\
+% ~             & \^{}гына/гына\tag{<postadv>\$}\^{}./.\tag{<sent>\$} \\
+{\bf Mor.\ analysis} & \^{}Ол/ол\tag{<det>}\tag{<dem>}/ол\tag{<prn>}\tag{<dem>}\tag{<nom>}/\textbf{ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}}\tag{\$} \\
+                     & \^{}енді/ен\tag{<n>}\tag{<acc>}/ен\tag{<v>}\tag{<iv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/ен\tag{<v>}\tag{<iv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/\textbf{енді\tag{<adv>}}\tag{\$}  \\
+							& \^ол/\textbf{ол\tag{<det>}\tag{<dem>}}/ол\tag{<prn>}\tag{<dem>}\tag{<nom>}/ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \\
+                     & \^дыбысты/\textbf{дыбыс\tag{<n>}\tag{<acc>}}\tag{\$} \\
+							& \^анығырақ/анық\tag{<adj>}\tag{<comp>}/\textbf{анық\tag{<adj>}\tag{<comp>}\tag{<advl>}}/анық\tag{<adj>}\tag{<comp>}\tag{<subst>}\tag{<nom>}\tag{\$} \\
+                     & \^ести/\textbf{есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}}\tag{\$}  \\
+                     & \^бастады/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/\textbf{баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}}\tag{\$}  \\
 \midrule

### 2013-04-18 paper/paper.tex
to get ^ you need all 4 characters: \^{} .  silly LaTeX

@@
                      & \^{}енді/ен\tag{<n>}\tag{<acc>}/ен\tag{<v>}\tag{<iv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/ен\tag{<v>}\tag{<iv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/\textbf{енді\tag{<adv>}}\tag{\$}  \\
-							& \^ол/\textbf{ол\tag{<det>}\tag{<dem>}}/ол\tag{<prn>}\tag{<dem>}\tag{<nom>}/ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \\
-                     & \^дыбысты/\textbf{дыбыс\tag{<n>}\tag{<acc>}}\tag{\$} \\
-							& \^анығырақ/анық\tag{<adj>}\tag{<comp>}/\textbf{анық\tag{<adj>}\tag{<comp>}\tag{<advl>}}/анық\tag{<adj>}\tag{<comp>}\tag{<subst>}\tag{<nom>}\tag{\$} \\
-                     & \^ести/\textbf{есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}}\tag{\$}  \\
-                     & \^бастады/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/\textbf{баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}}\tag{\$}  \\
+							& \^{}ол/\textbf{ол\tag{<det>}\tag{<dem>}}/ол\tag{<prn>}\tag{<dem>}\tag{<nom>}/ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \\
+                     & \^{}дыбысты/\textbf{дыбыс\tag{<n>}\tag{<acc>}}\tag{\$} \\
+							& \^{}анығырақ/анық\tag{<adj>}\tag{<comp>}/\textbf{анық\tag{<adj>}\tag{<comp>}\tag{<advl>}}/анық\tag{<adj>}\tag{<comp>}\tag{<subst>}\tag{<nom>}\tag{\$} \\
+                     & \^{}ести/\textbf{есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}}\tag{\$}  \\
+                     & \^{}бастады/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/\textbf{баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}}\tag{\$}  \\
 \midrule

### 2013-04-18 paper/paper.tex
Morphological disambiguation part of table

@@
 \midrule
-{\bf Mor. disambiguation}& \^{}Һава\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
- ~                  & \^{}җылы\tag{<adj>\$} \^{}гына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
+%{\bf Mor. disambiguation}& \^{}Һава\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
+{\bf Mor.\ disambiguation} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}\tag{\$} \^{}ол\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}\tag{\$} \\
+                     & \^{}анық\tag{<adj>}\tag{<comp>}\tag{<advl>}\tag{\$} \^{}есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}\tag{\$} \^{}баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}\tag{\$}\^{}.\tag{<sent>}\tag{\$} \\
 \midrule

### 2013-04-19 paper/paper.tex
мен үйге қайтуым керек <-> миңа өйгә кайтырга кирәк; add a note

@@
 Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').
+% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
+% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
+% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.
 
@@
 %	\begin{xlist}
-%		\ex Мен үйге баруым керек
-%		\gll мен үй-GA бар-у-Iм керек \\
+%		\ex Мен үйге қайтуым керек
+%		\gll мен үй-GA қайт-у-Iм керек \\
 %		I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need \\
@@
 %
-%		\ex Мин өйгә барырга кирәк
-%		\gll мин өй-GA бар-IргA кирәк \\
-%		I home-\gmk{Dat} go-\gmk{Inf} need \\
+%		\ex Миңа өйгә кайтырга кирәк
+%		\gll мин-GA өй-GA кайт-IргA кирәк \\
+%		I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need \\
 %		\glt \eng{I need to go home}
@@
 \begingl
-\gla Мен үйге баруым керек. //
-\glb мен үй-GA бар-у-Iм керек //
+\gla Мен үйге қайтуым керек. //
+\glb мен үй-GA қайт-у-Iм керек //
 \glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
@@
 \begingl
-\gla Мин өйгә барырга кирәк. //
-\glb мин өй-GA бар-IргA кирәк //
-\glb I home-\gmk{Dat} go-\gmk{Inf} need //
+\gla Миңа өйгә кайтырга кирәк. //
+\glb мин-GA өй-GA кайт-IргA кирәк //
+\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
 \glft \eng{I need to go home.} //
@@
 % FIXME: the following sh... stuff needs to be rephrased
-The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand.
-Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - "автомобиль" etc).
+The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand (translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
+Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - such as "автомобиль", "гонорар" etc).
 

### 2013-04-19 paper/paper.tex
example entries from the bilingual lexicon

@@
     <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
+    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e>
+    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
 \end{texttt}

### 2013-04-19 paper/paper.tex
selimcan forgot a newline

--- 2013-04-19 paper/paper.tex
@@
     <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
+    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
+    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
 \end{texttt}

### 2013-04-19 paper/paper.tex
some formatting improvements of figure 2

--- 2013-04-19 paper/paper.tex
@@
 \begin{figure*}[htbp]
-\begin{texttt}
+\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
+%\begin{texttt}
     <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{корал}<s n="n"/></r></p></e> \\
     <e><p><l>\textbf{құрал}<s n="n"/></l><r>\textbf{чара}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{борын}<s n="n"/></l><r>\textbf{морон}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{алы{\qipa ҫ}лы{\qipa ҡ}}<s n="n"/></r></p></e> \\
-    <e><p><l>\textbf{ераклык}<s n="n"/></l><r>\textbf{йыра{\qipa ҡ}лы{\qipa ҡ}}<s n="n"/></r></p></e>
-\end{texttt}
+    <e><p><l>\textbf{есім}<s n="n"/></l><r>\textbf{исем}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{туп}<s n="n"/></r></p></e> \\
+    <e><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{төркем}<s n="n"/></r></p></e> \\
+    <e r="RL"><p><l>\textbf{топ}<s n="n"/></l><r>\textbf{группа}<s n="n"/></r></p></e>
+}}%\end{texttt}
 \caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}

### 2013-04-20 paper/paper.tex
minor
a little work on bilingual lexicon section

@@
 % FIXME: the following sh... stuff needs to be rephrased
-The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand (translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
-Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - such as "автомобиль", "гонорар" etc).
+The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
+Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as "автомобиль", "гонорар" etc) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
 
-Entries consist largely of one-to-one stem-to-stem correspondences with part of speech, but also
-include some entries with ambiguous translations (see e.g., Fig.~\ref{fig:bidix}).
+Entries consist largely of one-to-one stem-to-stem correspondences with part of speech, but also include some entries with ambiguous translations (see e.g., Fig.~\ref{fig:bidix}).
 
@@
 
-The grammar currently has only four rules, but given the closeness of the languages, the 
-majority of ambiguity may be passed through from one language to the other.
+The grammar currently has fifty-eight rules.
+%but given the closeness of the languages, the 
+%majority of ambiguity may be passed through from one language to the other.
 

### 2013-04-20 paper/paper.tex
some disambig section additions
a pararaph in cg section

@@
 
-% FIXME: the following sh... stuff needs to be rephrased
 The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
@@
 The system has a morphological disambiguation module in the form of a 
-Constraint Grammar (CG) \citep{karlsson95}. The version of the formalism used is 
-vislcg3.\footnote{\url{http://beta.visl.sdu.dk/constraint_grammar.html}}
+Constraint Grammar (CG) \citep{karlsson95}. The version of the formalism used is vislcg3.\footnote{\url{http://beta.visl.sdu.dk/constraint_grammar.html}}
 
-The grammar currently has fifty-eight rules.
+%FIXME: Describe the gist of the rules, maybe with a few concrete examples.
+The grammar currently has 58 rules.  The goal of these rules is to select the correct analysis when there are multiple morphological analyses of Kazakh forms.
+%FIXME: It might be good to put some numbers about the morphological ambiguity of Kazakh
+
+The output of the morphological analyzer is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
+verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.
+
+We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
+adjective], [finite form], [substantivized verbal adjective, i.e. if the word it determines was ommitted], whereas many grammars
+do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.
+
+This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
+to another and most likely won't lead to translation errors. Disambiguating it, and, prior to that, having that kind of
+informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.
+
+This is the reason for a high ambiguity rate of the transducers.
+
 %but given the closeness of the languages, the 
@@
 
-We would like to thank the anonymous reviewers for their helpful comments in improving the paper.
-This work has been partially funded by Spanish Ministerio de Ciencia e Innovación through project TIN2009-14009-C02-01.
+This work has been partially funded by the Google Summer of Code and Google Code-in programs.
 

### 2013-04-21 paper/paper.tex
table 3 done (?)

@@
 \toprule
-{\bf (Kazakh) Input} & Ол енді ол дыбысты анығырақ ести бастады \\
+{\bf (Kazakh) Input} & Ол енді ол дыбысты анығырақ ести бастады. \\
 %{\bf (Tatar) Input} & Һава бүген бик әйбәт, җылы гына. \\ 
@@
                      & \^{}ести/\textbf{есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}}\tag{\$}  \\
-                     & \^{}бастады/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/\textbf{баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}}\tag{\$}  \\
+                     & \^{}бастады/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<pl>}/баста\tag{<v>}\tag{<tv>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<pl>} \\
+                     & \hspace{2em}/\textbf{баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}}\tag{\$}  \\
 \midrule
@@
 \midrule
-{\bf Lex. transfer} & \^{}Һава\tag{<n>}\tag{<nom>}/Һауа\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>}/бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>}/бик\tag{<adv>\$} \\
-( + selection)                     & \^{}әйбәт\tag{<adj>}/әйбәт\tag{<adj>\$}\^{},\tag{<cm>}/,\tag{<cm>\$} \^{}җылы\tag{<adj>}/йылы\tag{<adj>\$} \\ 
-~                     & \^{}гына\tag{<postadv>}/ғына\tag{<postadv>\$}\^{}.\tag{<sent>}/.\tag{<sent>\$}\\
+{\bf Lex. transfer} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<nom>}/Ул\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}/инде\tag{<adv>}/хәзер\tag{<adv>}\tag{\$} \\
+ ( + selection)     & \^{}ол\tag{<det>}\tag{<dem>}/ул\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}/тавыш\tag{<n>}\tag{<acc>}\tag{\$} \\
+                    & \^{}анық\tag{<adj>}\tag{<comp>}\tag{<advl>}/анык\tag{<adj>}\tag{<comp>}\tag{<advl>}\tag{\$}  \\
+						  & \^{}есті\tag{<v>}\tag{<tv>}<prc\_impf>/ишет\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}\tag{\$} \\
+                    & \^{}баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}/башла\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}\tag{\$}\^{}.\tag{<sent>}/.\tag{<sent>}\tag{\$} \\
+%\^{}Һава\tag{<n>}\tag{<nom>}/Һауа\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>}/бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>}/бик\tag{<adv>\$} \\
+%                     & \^{}әйбәт\tag{<adj>}/әйбәт\tag{<adj>\$}\^{},\tag{<cm>}/,\tag{<cm>\$} \^{}җылы\tag{<adj>}/йылы\tag{<adj>\$} \\ 
+%~                     & \^{}гына\tag{<postadv>}/ғына\tag{<postadv>\$}\^{}.\tag{<sent>}/.\tag{<sent>\$}\\
 \midrule
-{\bf Struct. transfer}& \^{}Һауа\tag{<n>}\tag{<nom>\$} \^{}бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
-~                    & \^{}йылы\tag{<adj>\$} \^{}ғына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
+{\bf Struct. transfer}& \^{}Ул\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<nom>}\tag{\$} \^{}инде\tag{<adv>}\tag{\$} \^{}ул\tag{<det>}\tag{<dem>}\tag{\$} \^{}тавыш\tag{<n>}\tag{<acc>}\tag{\$} \\
+                      & \^{}анык\tag{<adj>}\tag{<comp>}\tag{<advl>}\tag{\$} \^{}ишет\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}\tag{\$} \^{}башла\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}\tag{\$}\^{}.\tag{<sent>}\tag{\$} \\
+
+%\^{}Һауа\tag{<n>}\tag{<nom>\$} \^{}бөгөн\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
+%~                    & \^{}йылы\tag{<adj>\$} \^{}ғына\tag{<postadv>\$}\^{}.\tag{<sent>\$}\\
 \midrule
-{\bf Mor. generation} & Һауа бөгөн бик әйбәт, йылы ғына. \\
+{\bf Mor. generation} & Ул инде ул тавышны аныграк ишетә башлады. \\
 \bottomrule
 \end{tabular}
- \caption{Translation process for the phrase \emph{Һава бүген бик әйбәт, җылы гына} `The weather today is very nice, it is very warm'.}
+ \caption{Translation process for the phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
  %\vspace{-1em}
@@
 
-This work has been partially funded by the Google Summer of Code and Google Code-in programs.
+The work on this Kazakh-Tatar machine translation system was partially funded by the Google Summer of Code and Google Code-In programs.
 

### 2013-04-21 paper/paper.tex
numbers!

@@
 \setmonofont[Scale=0.7]{DejaVu Sans Mono}
+\newfontfamily\smallertt[Scale=0.55]{DejaVu Sans Mono}
 
@@
 The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
-Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as "автомобиль", "гонорар" etc) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
+Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
 
@@
 
+%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
 The output of the morphological analyzer is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
@@
 We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
-adjective], [finite form], [substantivized verbal adjective, i.e. if the word it determines was ommitted], whereas many grammars
+adjective], [finite form], [substantivized verbal adjective, i.e., if the word it determines was ommitted], whereas many grammars
 do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.
@@
 
-Likewise, lexical selection is not a large problem between Tatar and Bashkir, but a 
-number of rules can be written for ambiguous words; for example, the Tatar 
-word \emph{борын} `nose (person), nose (ship)' can be translated into Bashkir 
-as either \emph{танау} `nose (person)' or \emph{морон} `nose (ship)'. A lexical selection
-rule chooses the translation \emph{танау} if the immediate context includes a proper 
-name.
+While many lexical items have a similar range of meaning, lexical selection can sometimes be problematic between Kazakh and Tatar.
 
-Another example is the word \emph{катлаулы} `layered'.  It is always translated to Bashkir 
-as \emph{{\qipa ҡ}атмарлы}, except in the collocaton \emph{катлаулы мәсьәлә} `difficult matter/problem', which is translated as \emph{{\qipa ҡ}атлаулы мәсьәлә}.
+For example (see figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all meanings corresponding to its Tatar cognate корал; however, it is also used in the compound ақпарат құралдары \eng{mass media} (literally, \eng{means of information}), which translates to Tatar as мәгълүмат чаралары (which has the same literal translation).
+Hence, the Kazakh word құрал must have two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds to Tatar чара.
+A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation.
+%FIXME: Ilnar, we probably want an example of this rule
+
+Likewise, the Kazakh word топ has two translations into Tatar.
+One translation is туп \eng{ball} (which can also be доп in Kazakh), and the other translation is төркем \eng{group}.
+The bilingual dictionary also has the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated to Kazakh топ (i.e., analysed), but is never generated.
 
@@
 
-Lexical coverage of the system is calculated over a freely available corpus of Bashkir, the Bashkir
-Wikipedia,\footnote{\url{http://ba.wikipedia.org/}; {\tt bawiki-20111210-pages-articles.xml.bz2}} and over two freely available corpora of 
-Tatar, the Tatar Wikipedia\footnote{\url{http://tt.wikipedia.org/}; {\tt ttwiki-20111215-pages-articles.xml.bz2}} and the New Testament in Tatar.
-The version of the translation tested was {\tt {\small r37137}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}
+Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar.
+For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty's Kazakh-language service\footnote{\url{http://www.azattyq.org/}}, as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
+For Tatar, the Tatar Wikipedia\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}}, the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service \footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012.
+The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the coverage as calculated for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
 
 \begin{table}[htbp]
-  \begin{center}
-  \begin{tabular}{lrr}
+	\centering
+	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
+		\begin{tabular}{llrrr}
 			\toprule
-   Corpus                  & Tokens    & Coverage\\
+			Corpus          & Tokens & Coverage & stdev\\
 			\midrule
-   Tatar New Test.         & 163,603   & 72.04\% \\
-   Tatar Wikipedia         & 37,123    & 70.19\% \\
-   \midrule
-   Bashkir Wikipedia       & 12,267    & 65.99\% \\
+			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
+			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
+			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
 			\bottomrule
 		\end{tabular}
-    \caption{Na\"ive vocabulary coverage over the three corpora.}
- %\vspace{-1.5em}
+	}\\
+	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
+		\begin{tabular}{llrrr}
+			\toprule
+			Corpus          & Tokens & Coverage & stdev\\
+			\midrule
+			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
+			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
+			New Testament   & 137K   & 92.54\%  & 1.31\% \\
+			\bottomrule
+		\end{tabular}
+	}
+	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
 	\label{table:coverage}
-  \end{center}
 \end{table}
 
+%FIXME: replace old text
 As shown in Table~\ref{table:coverage}, the coverage is still far too low to be of use as a 
@@
 
+\begin{table}[htbp]
+	\centering
+	\begin{tabular}{llrrr}
+		\toprule
+		Corpus          & Tokens & Coverage & stdev\\
+		\midrule
+		RFERL 2010      & 3.2M   & 90.19\% & 0.23\% \\
+		RFERL 2012      & 2.9M   & 89.74\% & 0.59\% \\
+		Wikipedia       & 1.2M   & 80.75\% & 5.23\% \\
+		\bottomrule
+	\end{tabular}
+	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
+	\label{table:trimmedcoverage}
+\end{table}
+
+As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT system is comparable to that of the individual transducers, despite the fact that the vanilla transducers include more entries than the trimmed versions used in the translation system.
+
+%FIXME: describe new stuff
 To get an idea of the kind of performance that could be expected from the system, we 

### 2013-04-21 paper/paper.tex
add an example of a lexical selection rule
some small additions to new lrx stuff

@@
 Hence, the Kazakh word құрал must have two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds to Tatar чара.
-A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation.
-%FIXME: Ilnar, we probably want an example of this rule
+A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in figure~\ref{fig:lrx}.
+
+\begin{figure*}[htbp]
+\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
+%\begin{texttt}
+  <rule> \\
+    <match lemma="ақпарат" tags="n.attr"/> \\
+    <match lemma="құрал" tags="n.*.px3sp.*"><select lemma="чара"/></match> \\
+  </rule>
+}}%\end{texttt}
+\caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
+\label{fig:lrx}
+%\vspace{-1em}
+\end{figure*}
 
@@
 The bilingual dictionary also has the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated to Kazakh топ (i.e., analysed), but is never generated.
+
+The system currently has a total of 33 lexical selction rules.
 

### 2013-04-21 paper/paper.tex
some scribbles on evaluation

@@
 
-%FIXME: describe new stuff
-To get an idea of the kind of performance that could be expected from the system, we 
-translated a simple story from Tatar to Bashkir and vice versa. The story may be found 
-online,\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/xupaixkar/rasskaz}}
-and was used for pedagogical purposes in a recently workshop on MT
-for the languages of Russia.
+To measure the performance of the translator we used the Word Error Rate metiric -- an edit metric based on the
+Levenshtein distance \citep{levenshtein/1966}.
+
+We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
+The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
+Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
+on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
+morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
+the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
+Hence we can call it the development corpus. It presents an upper bound on the current performance
+of the system. Table FIXME http://pastebin.com/TaaWxJzN presents the WER for it.
+
+The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
+Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
+The WER results for it can be seen in the table FIXME http://pastebin.com/7tzGCgMX.
+
+Part of the corrections while postediting were due the lack of morphophonoligical rules in Tatar transducer,
+and were not translation errors as such.
+
+%The WER on the translation of the development corpus and its postedited version is twice as low.
+%Which, considering the relatively low number of rules, indicates that a high quality translation can be achived
+%by advancing these two modules.
+
+%To get an idea of the kind of performance that could be expected from the system, we 
+%translated a simple story from Tatar to Bashkir and vice versa. The story may be found 
+%online,\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/xupaixkar/rasskaz}}
+%and was used for pedagogical purposes in a recently workshop on MT
+%for the languages of Russia.
 
@@
 
-Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
-distance \citep{levenshtein/1966}. This measure was calculated once all the stems in the 
-text had been added to the system, thus presents an upper bound on the current performance
-of the transfer lexicon, and the disambiguation and transfer rules. The difference in 
-the number of unknown words between translating Tatar$\rightarrow$Bashkir and vice versa
-is because certain forms were not found due to lack of corresponding morphophonological rules.
+%Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
+%distance \citep{levenshtein/1966}. This measure was calculated once all the stems in the 
+%text had been added to the system, thus presents an upper bound on the current performance
+%of the transfer lexicon, and the disambiguation and transfer rules.
 
-We calculate the WER instead of other MT evaluation metrics such as BLEU as the WER is 
-geared towards a particular task, that of measuring postedition effort. The translations 
-of the story into Tatar and Bashkir were done in parallel to make them as close as possible,
-so using BLEU would give an over-optimistic view of the quality.
+%We calculate the WER instead of other MT evaluation metrics such as BLEU as the WER is 
+%geared towards a particular task, that of measuring postedition effort. The translations 
+%of the story into Tatar and Bashkir were done in parallel to make them as close as possible,
+%so using BLEU would give an over-optimistic view of the quality.
 
@@
 
-The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
-\begin{itemise}
-  \item Combinations of case and possessive suffixes,
-  \item Orthographical representations of phonology,
-  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
-\end{itemise}
+%The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
+%\begin{itemise}
+%  \item Combinations of case and possessive suffixes,
+%  \item Orthographical representations of phonology,
+%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
+%\end{itemise}
 
@@
 
-To our knowledge we have presented the first ever MT system between Tatar and Bashkir, and the first ever MT system involving Bashkir.
-The system is available as free/open-source software under the GNU GPL and the
-whole system may be downloaded from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/nursery/apertium-tt-ba}}
+To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production level
+coverage, but relatively small number of rules
+
+The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
+from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
 

### 2013-04-21 paper/paper.tex
concluding remarks

@@
 
-To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production level
-coverage, but relatively small number of rules
+To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production-level
+coverage, but rather a prototype-level number of rules. Though the impact of this relatively low number of rules on the
+quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
+the a high-quality translation between morphologically rich agglutinative languages is possible.
+
+We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
+increase it to 95% on the corporas we have, the main work will be improving the quality of the translation by adding
+more rules, starting with a Constraint Grammar module.
+The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
+transfer systems between all the Turkic language pairs.  Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.
 
@@
 from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
-
-We plan to continue development on the pair; the main work will be expanding the dictionaries with new lists of stems, and providing bilingual correspondences.
-The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make transfer systems between all the Turkic language pairs.
-Related work is currently ongoing with Chuvash--Turkish and Turkish--Kyrgyz.
 

### 2013-04-22 paper/paper.tex
wer numbers

@@
 %maybe this should be moved to 'Future plans'
-Within the Apertium project, work on several Turkic-language MT systems has been started, with the ultimate goal of creating independent but compatible
-finite-state transducers for each language.
+%Within the Apertium project, work on several Turkic-language MT systems has been started, with the ultimate goal of creating independent but compatible
+%finite-state transducers for each language.
 %% The idea is that we started Kazakh-Tatar pair to be able to take into account
@@
 Hence we can call it the development corpus. It presents an upper bound on the current performance
-of the system. Table FIXME http://pastebin.com/TaaWxJzN presents the WER for it.
+of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN
 
@@
 Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
-The WER results for it can be seen in the table FIXME http://pastebin.com/7tzGCgMX.
+The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
 
@@
   \midrule
-   \multirow{2}{*}{story} & tt$\rightarrow$ba & 311     & 9  & 8.97\% \\
-                          & ba$\rightarrow$tt & 312     & 1  & 7.72\%  \\
+  Development corpus      & kaz$\rightarrow$tat & 2457    & 2       & 15.19\% \\
   \bottomrule
   \end{tabular}
-    \caption{Word error rate and over the small test corpus.}
+    \caption{Word error rate over the development corpus.}
  %\vspace{-2.5em}
-    \label{table:wer}
+    \label{table:wer-development}
+  \end{center}
+\end{table}
+
+\begin{table}[htbp]
+  \begin{center}
+  \begin{tabular}{ccrrr}
+  \toprule
+   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
+  \midrule
+  Evaluation corpus       & kaz$\rightarrow$tat & 2862    & 43      & 36.57\% \\
+  \bottomrule
+  \end{tabular}
+    \caption{Word error rate over the evaluation corpus.}
+ %\vspace{-2.5em}
+    \label{table:wer-evaluation}
   \end{center}

### 2013-04-22 paper/paper.tex
some updates

@@
 \usepackage{latexsym}
-\usepackage[small,bf]{caption}
+%\usepackage[small,bf]{caption}
+\usepackage[bf]{caption}
 \usepackage{xltxtra}
-
+\usepackage{times}
 \usepackage{fontspec}
@@
 
-The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyzer into Tatar).
+The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyser into Tatar).
 Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
@@
 %FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
-The output of the morphological analyzer is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
+The output of the morphological analyser is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
 verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.
@@
 
-\begin{table}[htbp]
-	\centering
-	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
-		\begin{tabular}{llrrr}
-			\toprule
-			Corpus          & Tokens & Coverage & stdev\\
-			\midrule
-			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
-			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
-			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
-			\bottomrule
-		\end{tabular}
-	}\\
-	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
-		\begin{tabular}{llrrr}
-			\toprule
-			Corpus          & Tokens & Coverage & stdev\\
-			\midrule
-			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
-			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
-			New Testament   & 137K   & 92.54\%  & 1.31\% \\
-			\bottomrule
-		\end{tabular}
-	}
-	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
-	\label{table:coverage}
-\end{table}
+%%
+%%\begin{table}[htbp]
+%%	\centering
+%%	\subfloat[Coverage of Kazakh transducer over Kazakh corpora]{
+%%		\begin{tabular}{llrrr}
+%%			\toprule
+%%			Corpus          & Tokens & Coverage & stdev\\
+%%			\midrule
+%%			RFERL 2010      & 3.2M   & 90.26\% & 0.24\% \\
+%%			RFERL 2012      & 2.9M   & 89.81\% & 0.59\% \\
+%%			Wikipedia       & 1.2M   & 80.56\% & 5.44\% \\
+%%			\bottomrule
+%%		\end{tabular}
+%%	}\\
+%%	\subfloat[Coverage of Tatar transducer over Tatar corpora]{
+%%		\begin{tabular}{llrrr}
+%%			\toprule
+%%			Corpus          & Tokens & Coverage & stdev\\
+%%			\midrule
+%%			RFERL 2007-2012 & 1.2M   & 82.56\%  & 3.24\% \\
+%%			Wikipedia       & 128K   & 82.23\%  & 1.58\% \\
+%%			New Testament   & 137K   & 92.54\%  & 1.31\% \\
+%%			\bottomrule
+%%		\end{tabular}
+%%	}
+%%	\caption{Naïve vocabulary coverage of the stand-alone transducers.}
+%%	\label{table:coverage}
+%%\end{table}
 
 %FIXME: replace old text
-As shown in Table~\ref{table:coverage}, the coverage is still far too low to be of use as a 
-general broad-domain MT system, but we hope that it shows that a good proportion of 
-the morphology of both languages is in place.
+As can be seen in Table~\ref{table:coverage}, the coverage for the news corpora approaches
+that of a broad-coverage MT system, with one word in ten unknown. These unknown words are 
+principally proper nouns. For the Wikipedia corpus, the coverage is substantially worse. This is 
+due to \textbf{FIXME: why?}
 
-\begin{table}[htbp]
+\begin{table}
 	\centering
@@
 		\midrule
-		RFERL 2010      & 3.2M   & 90.19\% & 0.23\% \\
-		RFERL 2012      & 2.9M   & 89.74\% & 0.59\% \\
-		Wikipedia       & 1.2M   & 80.75\% & 5.23\% \\
+		RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
+		RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
+		Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
 		\bottomrule
@@
 	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
-	\label{table:trimmedcoverage}
+	\label{table:coverage}
 \end{table}
@@
 
-To measure the performance of the translator we used the Word Error Rate metiric -- an edit metric based on the
+To measure the performance of the translator we used the Word Error Rate metiric --- an edit metric based on the
 Levenshtein distance \citep{levenshtein/1966}.
@@
 morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
-the system, and all the rules (Constraint Grammar,lexical selection and transfer) were based on this corpus.
+the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
 Hence we can call it the development corpus. It presents an upper bound on the current performance
@@
 
-The second corpus -- a collection of only Radio Free Europe / Radio Liberty's
-Kazakh-language service articles -- was used solely for evaluation. It had a similar size -- 2862 words.
+The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
+Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2862 words.
 The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
@@
 
-\begin{table}[htbp]
+\begin{table}
   \begin{center}
@@
   \toprule
-   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
+   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
   \midrule
-  Development corpus      & kaz$\rightarrow$tat & 2457    & 2       & 15.19\% \\
+  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
   \bottomrule
+  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \end{tabular}
-    \caption{Word error rate over the development corpus.}
+    \caption{Word error rate over two corpora. The \texttt{devel} `development' corpus was used during development 
+       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
+       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
  %\vspace{-2.5em}
     \label{table:wer-development}
-  \end{center}
-\end{table}
-
-\begin{table}[htbp]
-  \begin{center}
-  \begin{tabular}{ccrrr}
-  \toprule
-   Corpus                 & Direction           & Tokens  & Unknown & WER  \\
-  \midrule
-  Evaluation corpus       & kaz$\rightarrow$tat & 2862    & 43      & 36.57\% \\
-  \bottomrule
-  \end{tabular}
-    \caption{Word error rate over the evaluation corpus.}
- %\vspace{-2.5em}
-    \label{table:wer-evaluation}
   \end{center}

### 2013-04-22 paper/paper.tex
some cleaning up of some stuff

@@
 Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar.
-For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty's Kazakh-language service\footnote{\url{http://www.azattyq.org/}}, as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
-For Tatar, the Tatar Wikipedia\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}}, the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service \footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012.
+
+For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty's Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
+
+For Tatar, the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012.
 The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the coverage as calculated for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
@@
 
-%FIXME: replace old text
-As can be seen in Table~\ref{table:coverage}, the coverage for the news corpora approaches
-that of a broad-coverage MT system, with one word in ten unknown. These unknown words are 
-principally proper nouns. For the Wikipedia corpus, the coverage is substantially worse. This is 
-due to \textbf{FIXME: why?}
+As can be seen in table~\ref{table:coverage}, the coverage for the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
+These unknown words are principally proper nouns.
+For the Wikipedia corpus, the coverage is substantially worse.
+This is due to the fact that this corpus is ``dirtier'', as well as to the fact that it has contains many more proper nouns.
 
@@
 	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
-	\label{table:coverage}
+	\label{table:trimmedcoverage}
 \end{table}

### 2013-04-22 paper/paper.tex
more numbers

@@
 For Tatar, the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012.
-The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the coverage as calculated for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
+
+The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
 
@@
 
-As can be seen in table~\ref{table:coverage}, the coverage for the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
-These unknown words are principally proper nouns.
-For the Wikipedia corpus, the coverage is substantially worse.
-This is due to the fact that this corpus is ``dirtier'', as well as to the fact that it has contains many more proper nouns.
-
 \begin{table}
 	\centering
+	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
 		\begin{tabular}{llrrr}
@@
 		\end{tabular}
-	\caption{Naïve trimmed coverage of the Kazakh-Tatar system}
+	}\\
+	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
+		\begin{tabular}{llrrr}
+			\toprule
+			Corpus          & Tokens & Coverage & stdev\\
+			\midrule
+			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
+			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
+			Wikipedia       & 128K   &      \% & $\pm$     \% \\
+			\bottomrule
+		\end{tabular}
+	}
+
+	\caption{Naïve coverage of the Kazakh-Tatar system}
 	\label{table:trimmedcoverage}
@@
 
-As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT system is comparable to that of the individual transducers, despite the fact that the vanilla transducers include more entries than the trimmed versions used in the translation system.
+As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
+These unknown words are principally proper nouns.
+For the Wikipedia corpus, the coverage is substantially worse.
+This is due to the fact that this corpus is ``dirtier'', as well as to the fact that it has contains many more proper nouns.
 

### 2013-04-22 paper/paper.tex
all trimmed coverage numbers
60 rules, not 58 :P
section 3.4 done?  selimcan, spectie, please proofread 3.4

@@
 % Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
-There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, FIXME, and FIXME.
+There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well as :.
 
@@
 In Tatar, many of these phrases use an invariant infinitival form.
-Some examples are shown below:
+Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
 
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\a
+\a\label{ex:kerek}
 \begingl
@@
 
-> Tatar infinitive (Kazakh ger+nom/acc/dat, Kazakh -GAlI vadv)
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\a\label{ex:try}
+\begingl
+\gla Айгүл оны табуға әрекет жасайды. //
+\glb Айгүл о-NI тап-у-GA әрекет жаса-E-дI //
+\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Aor} //
+\glft \eng{Aygül tries to find him.} //
+\endgl
+\a
+\begingl
+\gla Айгөл аны табарга тырыша. //
+\glb Айгөл а-NI тап-AргA тырыш-E //
+\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
+\glft \eng{Aygöl is trying to find him.} //
+\endgl
+\xe
+
+As shown in (\ref{ex:GAlI}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.
+
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\label{ex:GAlI}
+\a
+\begingl
+\gla Мен сенімен сөйлескелі келдім. //
+\glb мен сен-і-мен сөйле-с-GAлI кел-DI-м. //
+\glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\glft \eng{I came to speak with you.} //
+\endgl
+\a
+\begingl
+\gla Мин син белән сөйлешергә килдем. //
+\glb мин син белән сөйле-ш-IргA кил-DI-м. //
+\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\glft \eng{I came to speak with you.} //
+\endgl
+\xe
+
+This example also demonstrates the correspondence of the Kazakh intstrumental case to the Tatar postposition белән \eng{with}.
 
@@
 %FIXME: Describe the gist of the rules, maybe with a few concrete examples.
-The grammar currently has 58 rules.  The goal of these rules is to select the correct analysis when there are multiple morphological analyses of Kazakh forms.
+The grammar currently has 60 rules.  The goal of these rules is to select the correct analysis when there are multiple morphological analyses of Kazakh forms.
 %FIXME: It might be good to put some numbers about the morphological ambiguity of Kazakh
@@
 			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
-			Wikipedia       & 128K   &      \% & $\pm$     \% \\
+			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
 			\bottomrule

### 2013-04-22 paper/paper.tex
some minor formatting fixes

@@
 			\toprule
-			 & -pl & +pl \\\midrule
-			-formal & сен & сендер \\
-			+formal & сіз & сіздер \\\bottomrule
+			 & [$-$pl] & [$+$pl] \\\midrule
+			{}[$-$formal] & сен & сендер \\
+			{}[$+$formal] & сіз & сіздер \\\bottomrule
 		\end{tabular}
@@
 			\toprule
-			 & -pl & +pl \\\midrule
-			-formal & син & сез \\
-			+formal & сез & сез \\\bottomrule
+			 & [$-$pl] & [$+$pl] \\\midrule
+			{}[$-$formal] & син & сез \\
+			{}[$+$formal] & сез & сез \\\bottomrule
 		\end{tabular}%}
@@
 			\toprule
-			 & [+recent] & [-recent] \\\midrule
-			{}[+reliable] & -DI- & -GAн- \\
-			{}[-reliable] & -Iп- & -GAн екен- \\ \bottomrule
+			 & [$+$recent] & [$-$recent] \\\midrule
+			{}[$+$reliable] & -DI- & -GAн- \\
+			{}[$-$reliable] & -Iп- & -GAн екен- \\ \bottomrule
 		\end{tabular}
@@
 			\toprule
-			 & [-newlyAcq'd] & [+newlyAcq'd] \\\midrule
-			{}[+reliable] & -DI- & \multirow{2}*{-GAн- икән}\\
-			{}[-reliable] & -GAн- & \\\bottomrule
+			 & [$-$new] & [$+$new] \\\midrule
+			{}[$+$reliable] & -DI- & \multirow{2}*{-GAн- икән}\\
+			{}[$-$reliable] & -GAн- & \\\bottomrule
 		\end{tabular}

### 2013-04-22 paper/paper.tex
some minor changes
s/Мин син белән/Мин синең белән/ in the example

@@
 			 & [$-$new] & [$+$new] \\\midrule
-			{}[$+$reliable] & -DI- & \multirow{2}*{-GAн- икән}\\
-			{}[$-$reliable] & -GAн- & \\\bottomrule
+			{}[$+$reliable] & -DI- & -GAн- икән \\
+			{}[$-$reliable] & -GAн- & --- \\\bottomrule
 		\end{tabular}
@@
 \begingl
-\gla Мин син белән сөйлешергә килдем. //
-\glb мин син белән сөйле-ш-IргA кил-DI-м. //
-\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\gla Мин синең белән сөйлешергә килдем. //
+\glb мин син-нIң белән сөйле-ш-IргA кил-DI-м. //
+\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
 \glft \eng{I came to speak with you.} //

### 2013-04-22 paper/paper.tex
headed to Manchu, but now selimcan and spectie should really check §3.4

@@
 %FIXME: use a noun in examples, not infinitive/gerund
-An example of a difference in verb valencies is with the expression corresponding to ``to like something'' in Kazakh and Tatar.
-In Kazakh, the verb ұна is used, e.g.
-бауырсақ маған ұнайды ``I like bawyrsaq'', where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the object ``bawyrsaq'' in English is the grammatical subject in Kazakh.
-Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, e.g.
-мин бавырсак яратам ``I like bawyrsaq'', where the first person pronoun is the grammatical subject and ``bawyrsaq'' is the grammatical direct object (with no accusative suffix since it is indefinite).
+An example of a difference in verb valencies is with the expression corresponding to ``to like to do something'' in Kazakh and Tatar.
+In Kazakh, the verb ұна is used, as shown in example (\ref{ex:like-kaz}), where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the gerundive ``writing dictionaries'' is the grammatical subject.
+Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, as shown in example (\ref{ex:like-tat}), where the first person pronoun is in nominative case as the grammatical subject and the infinitival verb phrase is the grammatical direct object.
 %(does this count as syntactic?)
+
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\a\label{ex:like-kaz}
+\begingl
+\gla Маған сөздік түзген ұнайды. //
+\glb маған сөздік түз-GAн ұна-E-дI //
+\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
+\glft \eng{I like writing dictionaries.} //
+\endgl
+\a\label{ex:like-tat}
+\begingl
+\gla Мин сүзлек төзергә яратам. //
+\glb мин сүзлек төз-IргA ярат-E-м //
+\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
+\glft \eng{I like writing dictionaries.} //
+\endgl
+\xe
 
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\a\label{ex:kerek}
+\label{ex:kerek}
+\a
 \begingl
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\a\label{ex:try}
+\label{ex:try}
+\a
 \begingl
@@
 \gla Мен сенімен сөйлескелі келдім. //
-\glb мен сен-і-мен сөйле-с-GAлI кел-DI-м. //
+\glb мен сен-I-мен сөйле-с-GAлI кел-DI-м. //
 \glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
@@
 \gla Мин синең белән сөйлешергә килдем. //
-\glb мин син-нIң белән сөйле-ш-IргA кил-DI-м. //
+\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
 \glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //

### 2013-04-22 paper/paper.tex
a minor fix to an example and to LAME NON-LOCALISED BIBTEX

@@
 \begingl
-\gla Айгүл оны табуға әрекет жасайды. //
-\glb Айгүл о-NI тап-у-GA әрекет жаса-E-дI //
-\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Aor} //
-\glft \eng{Aygül tries to find him.} //
+\gla Айгүл оны табуға әрекет жасап жүр. //
+\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
+\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
+\glft \eng{Aygül is trying to find him.} //
 \endgl

### 2013-04-22 paper/paper.tex
cooperate with spectie
updates and a coupe of otodos

@@
 
+\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}
+
 %\setlength\titlebox{6.5cm}    % You can expand the title box if you
@@
 This paper presents a bidirectional machine translation system between Kazakh and Tatar.
+\todo{expand abstract}
 \end{abstract}
@@
 
-Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish-Kyrgyz, Azeri-Turkish), but the Kazakh-Tatar system described by the present study is the closest to production-ready of them.
+Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Turkish-Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.
 Among these systems is a prototype Tatar-Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
 
-Besides these systems, several previous works on making machine translation systems between Turkic languages exist, although to our knowledge none are publically available.
-Some MT systems have been reported that translate between Turkish and other Turkic languages, including Turkish-Crimean Tatar \citep{altintas01},
-Turkish-Azerbaijani \citep{hamzaoglu93}, Turkish-Tatar \citep{suleymanov08}, and
-Turkish-Turkmen \citep{tantug07}, though none of these have been released to a public audience.
+Besides these systems, several previous works on making machine translation systems between Turkic languages 
+exist, although to our knowledge none are publically available.
+Some MT systems have been reported that translate between Turkish and other Turkic languages, 
+including Turkish--Crimean Tatar \citep{altintas01},
+Turkish--Azerbaijani \citep{hamzaoglu93}, Turkish--Tatar \citep{suleymanov08}, and
+Turkish--Turkmen \citep{tantug07}, though none of these have been released to a public audience. The Turkish--Azerbaijani
+pair is also available through Google Translate.\footnote{\url{http://translate.google.com}}
 
@@
 Both Tatar and Kazakh belong to the Kypchak (or Northwestern) group of Turkic
-languages.
-The spoken and written languages share some level of mutual intelligibility to naïve native speakers, though this is somewhat limited, and is obscured by different orthographical conventions and some opaque correspondences.
+languages.  The spoken and written languages share some level of mutual intelligibility to native 
+speakers, though this is somewhat limited, and is obscured by different orthographical 
+conventions and some opaque correspondences.
 
@@
 %a generalised summary, 2 or 3 small specific examples
-As closely related languages, Kazakh and Tatar share many phonological processes, including front-back vowel harmony systems, consonant voicing assimilation, and even a typologically rare consonantal nasal harmony system.
-However, the differing details of these processes and the existence of processes unique to each language render Kazakh and Tatar fairly different.
-For example, Kazakh has a ubiquitous system of desonorisation of the initial sonorants found in many common morphemes.
-Furthermore, Tatar has nasal assimilation of the initial /l/ of the plural-suffix.
+As closely related languages, Kazakh and Tatar share many phonological processes, including 
+front-back vowel harmony systems, consonant voicing assimilation, and even a typologically rare 
+consonantal nasal harmony system.  However, the differing details of these processes and the existence of 
+processes unique to each language render Kazakh and Tatar fairly different.  For example, Kazakh has a 
+ubiquitous system of desonorisation of the initial sonorants found in many common morphemes.  Furthermore, Tatar 
+has nasal assimilation of the initial /l/ of the plural-suffix.
 
@@
 %a generalised summary, 1 or 2 small specific examples
-The standard varieties of Kazakh and Tatar our system deals with are both written in Cyrillic, though their implementations of Cyrillic differ in many ways.
+The standard varieties of Kazakh and Tatar our system deals with are both written in Cyrillic, though their 
+implementations of Cyrillic differ in many ways.
 
-While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent vowels (and employing ``hard'' and ``soft signs'' when these fail) to differentiate the two, and Kazakh has two series of obstruents (e.g., ‹к› and ‹қ›).
+While Tatar and Kazakh both have a velar/uvular obstruent distinction (e.g., /k/ vs.\ /q/) that interacts with 
+adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent 
+vowels (and employing ъ `hard sign' and ь `soft sign' when these fail) to differentiate the two, and Kazakh 
+has two series of obstruents (e.g., ‹к› and ‹қ›).
 
-Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.
-The quality of these vowels is necessary to know in order to predict the quality of following harmonising vowels.
-%, so this convention presents accute challenges to producing an accurate morphological transducer.
-Additionally, Tatar and Kazakh both use ``yoticised'' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,
+Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before 
+glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both 
+written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.  The quality of these vowels is necessary to know in 
+order to predict the quality of following harmonising vowels. %, so this convention presents accute challenges to producing an accurate morphological transducer.
+Additionally, Tatar and Kazakh both use `yoticised' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in 
+Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,
 
-All of these orthographical conventions present accute challenges to designing accurate morphological transducers for the languages.
+All of these orthographical conventions present acute challenges to designing accurate morphological transducers for the languages.
 
 \subsection{Morphological differences}
-There are a number of examples where the morphologies of Kazakh and Tatar are rather different, including morphemes in one language that don't exist in the other, entirely different uses of the same morpheme combinations, and morphotactic differences (i.e., allowable ordering and placement of morphemes).
+There are a number of examples where the morphologies of Kazakh and Tatar are rather different, including 
+morphemes in one language that do not exist in the other, entirely different uses of the same morpheme 
+combinations, and morphotactic differences (i.e., allowable ordering and placement of morphemes).
 
-An example of a morpheme that doesn't exist in one of the two languages is Kazakh -\texttt{\{E\}т\{I\}н}, which is used to form non-past verbal adjectives and verbal nouns.
-The semantically equivalent structure in Tatar is -\texttt{\{E\} торган}, which historically corresponds to the source of the Kazakh morpheme; however, the use of -\texttt{\{E\} тұрған} in modern Kazakh is different from that of -\texttt{\{E\}т\{I\}н}.
+An example of a morpheme that does not exist in one of the two languages is Kazakh -\texttt{\{E\}т\{I\}н}, 
+which is used to form non-past verbal adjectives and verbal nouns.  The semantically equivalent structure 
+in Tatar is -\texttt{\{E\} торган}, which historically corresponds to the source of the Kazakh morpheme; 
+however, the use of -\texttt{\{E\} тұрған} in modern Kazakh is different from that of -\texttt{\{E\}т\{I\}н}.
 
@@
 This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
-Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a four-way distinction in Kazakh's 2nd person system (both pronouns and agreement suffixes), where Tatar only has a two-way distinction.
-Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}.
+Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a 
+four-way distinction in Kazakh's 2nd person system (both pronouns and agreement suffixes), where Tatar only 
+has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], 
+whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}.
 \begin{table}[htbp]
@@
 
-This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns are only used for emphasis and clarification.
-However, this difference between Tatar and Kazakh in the second person system runs much deeper than just the pronoun system.
-Since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.
+This systematic difference would seem to be a minor issue, since, as is typical in pro-drop languages, pronouns 
+are only used for emphasis and clarification.  However, this difference between Tatar and Kazakh in 
+the second-person system runs much deeper than just the pronoun system.  Since all finite verb forms 
+morphologically agree in person and number with their subject and all possessed nouns agree in person and 
+number with their possessor (even when there is no overt pronoun, in either situation), the Kazakh and Tatar 
+systems of agreement suffixes reflect the same pattern; i.e., there are several sets of agreement morphemes 
+which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of 
+suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.
 
-The past tense systems of Kazakh and Tatar have a many-to-many correspondence.
-As shown in table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.
-As an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a distant non-eyewitness past (\texttt{-GAн екен-}).
-On the other hand, these two non-eyewitness past forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
+The past tense systems of Kazakh and Tatar have a many-to-many correspondence.  As shown 
+in table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh 
+differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it 
+may be better expressed as simply ``reliability of knowledge'' (which indeed often equates to whether the 
+knowledge was acquired first-hand or not) in many cases.} (where [-eyewitness] is used for cases of both 
+potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only 
+three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.  As 
+an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness 
+past tense morpheme (\texttt{-GAн-}) while Kazakh has a recent non-eyewitness past (\texttt{-Iп-}) and a 
+distant non-eyewitness past (\texttt{-GAн екен-}).  On the other hand, these two non-eyewitness past 
+forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas 
+in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
 
@@
 
-Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.
-Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').
+Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh 
+distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
+different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although 
+in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in 
+Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
 % I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
@@
 
-Another morphotactic difference between Kazakh and Tatar is found with the negative forms of the cognate \texttt{-GAн-} past tenses.
-In Kazakh, the negative form of the non-recent reliable-information past tense is \texttt{-GAн емес-}, whereas in Tatar, the negative form of the non-eyewitness past tense is \texttt{-мAGAн-}.
+Another morphotactic difference between Kazakh and Tatar is found with the negative forms of the 
+cognate \texttt{-GAн-} past tenses.  In Kazakh, the negative form of the non-recent reliable-information past tense 
+is \texttt{-GAн емес-}, whereas in Tatar, the negative form of the non-eyewitness past tense is \texttt{-мAGAн-}.
 
@@
 % Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
-There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well as :.
+There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb 
+valencies in equivalent translations, as well as :.
 
@@
 %FIXME: use a noun in examples, not infinitive/gerund
-An example of a difference in verb valencies is with the expression corresponding to ``to like to do something'' in Kazakh and Tatar.
-In Kazakh, the verb ұна is used, as shown in example (\ref{ex:like-kaz}), where the subject ``I'' in English is expressed through a dative experiencer in Kazakh and the gerundive ``writing dictionaries'' is the grammatical subject.
-Tatar, on the other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, as shown in example (\ref{ex:like-tat}), where the first person pronoun is in nominative case as the grammatical subject and the infinitival verb phrase is the grammatical direct object.
+An example of a difference in verb valencies is with the expression corresponding 
+to ``to like to do something'' in Kazakh and Tatar.  In Kazakh, the verb ұна is used, as shown in 
+example (\ref{ex:like-kaz}), where the subject ``I'' in English is expressed through a dative 
+experiencer in Kazakh and the gerundive ``writing dictionaries'' is the grammatical subject.  Tatar, on the 
+other hand, uses a verb whose arguments correspond to the arguments of ``to like'' in English, as 
+shown in example (\ref{ex:like-tat}), where the first person pronoun is in nominative case as 
+the grammatical subject and the infinitival verb phrase is the grammatical direct object.
 %(does this count as syntactic?)
@@
 
-In Kazakh, a gerund (i.e., verbal noun) with case marking and sometimes person agreement in the form of possessive suffixes is used to make a verb phrase an argument to certain other main phrases.
-In Tatar, many of these phrases use an invariant infinitival form.
-Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
+In Kazakh, a gerund (i.e., verbal noun) with case marking and sometimes person agreement in the form of 
+possessive suffixes is used to make a verb phrase an argument to certain other main phrases.  In Tatar, many 
+of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
 
@@
 \gla Мен сенімен сөйлескелі келдім. //
-\glb мен сен-I-мен сөйле-с-GAлI кел-DI-м. //
-\glb I you-\gmk{Acc}-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\glb мен сен-мен сөйле-с-GAлI кел-DI-м. //
+\glb I you-\gmk{Inst} talk-\gmk{Coop}-\gmk{VAdv} come-\gmk{Ifi}-\gmk{1p.Sg} //
 \glft \eng{I came to speak with you.} //
@@
 
-The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly by hand (i.e., by translating Kazakh stems unrecognized by the morphological analyser into Tatar).
-Some toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia; also, some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
+The bilingual lexicon currently contains 9,269 stem-to-stem correspondences and was built mostly 
+by hand (i.e., by translating Kazakh stems unrecognised by the morphological analyser into Tatar).  Some 
+toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia \citep{tyers08}; also, 
+some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual 
+dictionary automatically by taking the intersection of Russian and Kazakh wordlists.
 
@@
 %FIXME: Describe the gist of the rules, maybe with a few concrete examples.
-The grammar currently has 60 rules.  The goal of these rules is to select the correct analysis when there are multiple morphological analyses of Kazakh forms.
+The grammar currently has 60 rules.  The goal of these rules is to select the correct analysis when there 
+are multiple morphological analyses of Kazakh forms.
 %FIXME: It might be good to put some numbers about the morphological ambiguity of Kazakh
@@
 %FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
-The output of the morphological analyser is highly ambigious. Both Kazakh and Tatar have a lot of affixes which join the
+The output of the morphological analyser is highly ambiguous. Both Kazakh and Tatar have a lot of affixes which join the
 verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.
@@
 We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
-adjective], [finite form], [substantivized verbal adjective, i.e., if the word it determines was ommitted], whereas many grammars
+adjective], [finite form], [substantivised verbal adjective, i.e., if the word it determines was omitted], whereas many grammars
 do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.
@@
 This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
-to another and most likely won't lead to translation errors. Disambiguating it, and, prior to that, having that kind of
+to another and most likely will not lead to translation errors. Disambiguating it, and, prior to that, having that kind of
 informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.
@@
 
-For example (see figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all meanings corresponding to its Tatar cognate корал; however, it is also used in the compound ақпарат құралдары \eng{mass media} (literally, \eng{means of information}), which translates to Tatar as мәгълүмат чаралары (which has the same literal translation).
-Hence, the Kazakh word құрал must have two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds to Tatar чара.
+For example (see figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
+meanings corresponding to its Tatar cognate корал; however, it is also used in the 
+compound ақпарат құралдары \eng{mass media} (literally, \eng{means of information}), which translates 
+to Tatar as мәгълүмат чаралары (which has the same literal translation).  Hence, the Kazakh word құрал must have 
+two entries in the bilingual lexicon: one that corresponds to Tatar корал and one that corresponds 
+to Tatar чара.
 A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in figure~\ref{fig:lrx}.
@@
 
-Likewise, the Kazakh word топ has two translations into Tatar.
-One translation is туп \eng{ball} (which can also be доп in Kazakh), and the other translation is төркем \eng{group}.
-The bilingual dictionary also has the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated to Kazakh топ (i.e., analysed), but is never generated.
+Likewise, the Kazakh word топ has two translations into Tatar.  One translation is туп \eng{ball} (which can also 
+be доп in Kazakh), and the other translation is төркем \eng{group}.  The bilingual dictionary also has 
+the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated 
+to Kazakh топ (i.e., analysed), but is never generated.
 
@@
 \end{tabular}
- \caption{Translation process for the phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
+ \caption{Translation process for the Kazakh phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
  %\vspace{-1em}
@@
 
-The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
+%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
 
@@
 
-As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
-These unknown words are principally proper nouns.
-For the Wikipedia corpus, the coverage is substantially worse.
-This is due to the fact that this corpus is ``dirtier'', as well as to the fact that it has contains many more proper nouns.
+As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
+system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over 
+the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
+For the Wikipedia corpus, the coverage 
+is substantially worse.  This is due to the fact that this corpus is ``dirtier'' --- 
+containing more orthographical errors, \emph{wiki} code, and repetitions, as well as to the 
+fact that it has contains many more proper nouns.
 
-To measure the performance of the translator we used the Word Error Rate metiric --- an edit metric based on the
+To measure the performance of the translator we used the Word Error Rate metric --- an edit-distance metric based on the
 Levenshtein distance \citep{levenshtein/1966}.
 
-We had two small Kazakh corporas alongside with their postedited translations into Tatar to measure the WER.
-The first one (2457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
+We had two small Kazakh corpora along with their postedited translations into Tatar to measure the WER.
+The first one (2,457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
 Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
@@
 The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
-Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2862 words.
+Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2,862 words.
 The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
@@
 and were not translation errors as such.
+
+\todo{Example of lack of phon rule}
 
@@
 quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
-the a high-quality translation between morphologically rich agglutinative languages is possible.
+the a high-quality translation between morphologically-rich agglutinative languages is possible.
 
 We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
-increase it to 95% on the corporas we have, the main work will be improving the quality of the translation by adding
-more rules, starting with a Constraint Grammar module.
+increase it to 95\% on the corpora we have, the main work will be improving the quality of the translation by adding
+more rules, starting with the Constraint Grammar module.
 The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make
@@
 
-The work on this Kazakh-Tatar machine translation system was partially funded by the Google Summer of Code and Google Code-In programs.
+The work on this Kazakh-Tatar machine translation system was partially funded by the Google Summer of Code and Google Code-In programmes.
 

### 2013-04-22 paper/paper.tex
first go of expected output vs current output examples

@@
 possessive suffixes is used to make a verb phrase an argument to certain other main phrases.  In Tatar, many 
-of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
+of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try})
+(where a is the original Kazakh sentence, b - the current output of the translator and c - the expected output.
 
@@
 \a
+\begingl
+\gla Мин син белән сөйлешергә килдем. //
+\glb мин син белән сөйле-ш-IргA кил-DI-м. //
+\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\glft \eng{I came to speak with you.} //
+\endgl
 \begingl

### 2013-04-22 paper/paper.tex
a few additions

@@
 \newfontfamily\smallertt[Scale=0.55]{DejaVu Sans Mono}
+\newfontfamily\smallerrm[Scale=0.85]{Times New Roman}
 
@@
 \begin{table}[htbp]
+	\centering
 %\begin{subtables}
@@
 			\toprule
-			 & [$-$pl] & [$+$pl] \\\midrule
-			{}[$-$formal] & сен & сендер \\
-			{}[$+$formal] & сіз & сіздер \\\bottomrule
+			 & {[$-$pl]} & {[$+$pl]} \\\midrule
+			{[$-$formal]} & сен & сендер \\
+			{[$+$formal]} & сіз & сіздер \\\bottomrule
 		\end{tabular}
@@
 	%}\subtable[Tatar 2nd person pronouns]{
-	}\vspace{1.5ex}
+	}\\
 	\subfloat[Tatar 2nd person pronouns]{
@@
 			\toprule
-			 & [$-$pl] & [$+$pl] \\\midrule
-			{}[$-$formal] & син & сез \\
-			{}[$+$formal] & сез & сез \\\bottomrule
+			 & {[$-$pl]} & {[$+$pl]} \\\midrule
+			{[$-$formal]} & син & сез \\
+			{[$+$formal]} & сез & сез \\\bottomrule
 		\end{tabular}%}
@@
 % Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
-There are a number of syntactic differences in Tatar and Kazakh, which include differences in verb 
-valencies in equivalent translations, as well as :.
+There are a number of minor syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well Tatar's reliance on a ``true'' infinitive that is used in place of various verbal noun and verb adverb forms in Kazakh.
 
@@
 
-This example also demonstrates the correspondence of the Kazakh intstrumental case to the Tatar postposition белән \eng{with}.
+This example also demonstrates the correspondence of the Kazakh intstrumental case -Мен to the Tatar postposition белән \eng{with}, which are cognate structures; while their phonology and orthographic standards differ, they are largely parallel in use.
 
@@
 
-The second corpus --- a collection of only Radio Free Europe / Radio Liberty's
-Kazakh-language service articles --- was used solely for evaluation. It had a similar size --- 2,862 words.
-The WER results for it can be seen in the table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
+The second corpus --- a collection of articles from Radio Free Europe / Radio Liberty's Kazakh-language service --- was used solely for evaluation.  It had a similar size --- 2,862 words.
+The WER results for this corpus are presented in table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
 
-Part of the corrections while postediting were due the lack of morphophonoligical rules in Tatar transducer,
-and were not translation errors as such.
+Some of the corrections that were made as part of postediting were due the lack of morphophonoligical rules in the Tatar transducer, and were not translation errors as such.
 

### 2013-04-22 paper/paper.tex
a space fix

@@
 toponyms and other proper names were translated semi-automatically by looking up links in Wikipedia \citep{tyers08}; also, 
-some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.) were added to the bilingual 
+some Russian loanwords common to both languages (such as автомобиль, гонорар, etc.)\ were added to the bilingual 
 dictionary automatically by taking the intersection of Russian and Kazakh wordlists.

### 2013-04-22 paper/paper.tex
nomoreneeded
firespeaker forgot to comment this out

@@
 possessive suffixes is used to make a verb phrase an argument to certain other main phrases.  In Tatar, many 
-of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try})
-(where a is the original Kazakh sentence, b - the current output of the translator and c - the expected output.
+of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
 
@@
 \endgl
+%\a
+%\begingl
+%\gla Мин син белән сөйлешергә килдем. //
+%\glb мин син белән сөйле-ш-IргA кил-DI-м. //
+%\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
+%\glft \eng{I came to speak with you.} //
+%\endgl
 \a
-\begingl
-\gla Мин син белән сөйлешергә килдем. //
-\glb мин син белән сөйле-ш-IргA кил-DI-м. //
-\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
-\glft \eng{I came to speak with you.} //
-\endgl
 \begingl

### 2013-04-22 paper/paper.tex
listings

@@
 \defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
+\usepackage{listings}
+
 \newfontfeature{IPA}{+mgrk}
@@
 %\DeclareFontShape{EU1}{TimesNewRoman(0)}{m}{sc}{<->ssub * TimesNewRoman(1)/m/sc}{}
+
+\lstset{language=XML, basicstyle=\ttfamily,
+    columns=fullflexible, keepspaces=true, morekeywords={lemma, tags, rule, match, or, select, remove, rules},
+    literate={á}{\'{a}}1 {é}{\'{e}}1 {í}{\'{\i}}1 {ó}{\'{o}}1 {ç}{\c{c}}1
+ {ñ}{\~{n}}1}
 
@@
 
-\begin{figure*}[htbp]
-\hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
-%\begin{texttt}
-  <rule> \\
-    <match lemma="ақпарат" tags="n.attr"/> \\
-    <match lemma="құрал" tags="n.*.px3sp.*"><select lemma="чара"/></match> \\
+\begin{figure}
+\begin{lstlisting}
+<rule> 
+  <match lemma="ақпарат" tags="n.attr"/> 
+  <match lemma="құрал" tags="n.*.px3sp.*">
+    <select lemma="чара"/>
+  </match> 
 </rule>
-}}%\end{texttt}
+\end{lstlisting}
+
+\todo{A couple more lexical-selection rules }
 \caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
@@
 %\vspace{-1em}
-\end{figure*}
+\end{figure}
 

### 2013-04-22 paper/paper.tex
maybe it will compile right now for you guys?
add my paper

@@
 \usepackage{fontspec}
-\defaultfontfeatures{PunctuationSpace=3,Scale=MatchLowercase,Mapping=tex-text}
+\defaultfontfeatures{Scale=MatchLowercase,Mapping=tex-text}
 \usepackage{listings}
@@
   return more than one TL LF for a single SL LF.
-\item A \emph{lexical selection} module which chooses, based on context 
+\item A \emph{lexical selection} module \citep{tyers12a} which chooses, based on context 
   rules the most adequate translation of ambiguous source language LFs.

### 2013-04-22 paper/paper.tex
add › to hard/soft sign

@@
 
-Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Turkish-Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.
-Among these systems is a prototype Tatar-Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
+Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Tatar--Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.
+Among these systems is a prototype Tatar--Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
 
@@
 adjacent vowels, the Tatar orthography only has one series of letters (e.g., ‹к›), relying on adjacent 
-vowels (and employing ъ `hard sign' and ь `soft sign' when these fail) to differentiate the two, and Kazakh 
+vowels (and employing ‹ъ› `hard sign' and ‹ь› `soft sign' when these fail) to differentiate the two, and Kazakh 
 has two series of obstruents (e.g., ‹к› and ‹қ›).

### 2013-04-22 paper/paper.tex
abstract

@@
 \begin{abstract}
-This paper presents a bidirectional machine translation system between Kazakh and Tatar.
+This paper presents a bidirectional machine translation system between Kazakh and Tatar, two Turkic languages.
+Background on the differences between the languages is presented, followed by how the system was designed to handle some of these differences.
+We provide an evaluation of the system's performance and directions for future work.
 \todo{expand abstract}

### 2013-04-22 paper/paper.tex
various fixes

@@
 
-The paper will be laid out as follows: Section~\ref{sec:lang}\ gives a brief
-description of the two languages; Section~\ref{sec:prev}\ gives a short review
-of some previous work in the area of Turkic--Turkic language translation;
+The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a brief
+description of the two languages; Section~\ref{sec:lang}\ gives a short review
+of some previous work in the area of Turkic--Turkic language translation and compares the grammar of Kazakh and Tatar;
 Section~\ref{sec:sys}\ describes the system and the tools used to construct it;
@@
 Besides these systems, several previous works on making machine translation systems between Turkic languages 
-exist, although to our knowledge none are publically available.
+exist, although to our knowledge none are publicly available except for the Turkish--Azerbaijani pair available through Google Translate.\footnote{\url{http://translate.google.com}}
 Some MT systems have been reported that translate between Turkish and other Turkic languages, 
@@
 Turkish--Azerbaijani \citep{hamzaoglu93}, Turkish--Tatar \citep{suleymanov08}, and
-Turkish--Turkmen \citep{tantug07}, though none of these have been released to a public audience. The Turkish--Azerbaijani
-pair is also available through Google Translate.\footnote{\url{http://translate.google.com}}
+Turkish--Turkmen \citep{tantug07}, though none of these have been released to a public audience.
 
@@
 
-For Tatar, the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012.
+For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.
 

### 2013-04-22 paper/paper.tex
put it back in with link to kaz-tat in staging

@@
 
-%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
+The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
 

### 2013-04-22 paper/paper.tex
one TODO down?  spectie, check

@@
 
-Some of the corrections that were made as part of postediting were due the lack of morphophonoligical rules in the Tatar transducer, and were not translation errors as such.
+Some of the corrections that were made as part of postediting were were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.
+A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.
+
+\begin{figure}[htbp]
+	\begin{tabular}{lll}
+		\toprule
+		form & current & expected \\\midrule
+		ди\tag{<v>}\tag{<tv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & дия & ди \\
+		йөр\tag{<v>}\tag{<iv>}\tag{<pres>}\tag{<p3>}\tag{<sg>} & йөрә & йөри \\
+		кияү\tag{<n>}\tag{<px3sp>}\tag{<nom>} & кияүе & кияве \\
+		укы\tag{<v>}\tag{<iv>}\tag{<ger>}\tag{<nom>} & укыу & уку \\\bottomrule
+	\end{tabular}
+	\caption{Examples of some phonological problems in the Tatar transducer.}
+	\label{tab:phonerrors}
+\end{figure}
 

### 2013-04-22 paper/paper.tex
aaaand centered
minor stuff
+1 lexical selection rule
another lrx rule

@@
 \end{lstlisting}
-
+\begin{lstlisting}
+  <rule>
+<!-- 61. Германия құрамасы Еуропа чемпионатын үш рет ұтып алған.
+     61. Германия җыелмасы Аурупа чемпионатын өч тапкыр откан. -->
+    <match tags="num"/>
+    <match lemma="рет" tags="n.*"><select lemma="тапкыр"/></match>
+  </rule>
+\end{lstlisting}
+\begin{lstlisting}
+  <rule>
+    <match lemma="рухани" tags="adj"/>
+    <match lemma="өмір" tags="n.*"><select lemma="тормыш"/></match>
+  </rule>
+\end{lstlisting}
 \todo{A couple more lexical-selection rules }
@@
 \begin{figure}[htbp]
+	\centering
 	\begin{tabular}{lll}
@@
   \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
+  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \bottomrule
-  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \end{tabular}

### 2013-04-22 paper/paper.tex
error analysis
new stuff cleaning blah

--- 2013-04-22 paper/paper.tex
@@
 
-\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}
+%\newcommand{\todo}[1]{\{\textbf{TODO: #1}\}}
+\newcommand{\todo}[1]{}
 
@@
 \begin{figure}
+%	\subfloat[A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.]{
 \begin{lstlisting}
@@
 \end{lstlisting}
-
+%}\\
+%\subfloat[]{
+%}
 \todo{A couple more lexical-selection rules }
+%\caption{Some lexical selection rules in Kazakh}
 \caption{A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.}
@@
 
-The system currently has a total of 33 lexical selction rules.
+Another lexical selection rule is shown in figure~\ref{fig:lrx2}.
+This rule ensures that the Kazakh word рет \eng{row, time(s)} is translated as тапкыр \eng{time(s)} when it is used with a number; otherwise it translates as рәт \eng{row}.
+\begin{figure}
+\begin{lstlisting}
+  <rule>
+    <match tags="num"/>
+    <match lemma="рет" tags="n.*">
+        <select lemma="тапкыр"/>
+    </match>
+  </rule>
+\end{lstlisting}
+\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
+\label{fig:lrx2}
+\end{figure}
+
+Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).
+The lexical selection rule is provided in figure~\ref{fig:lrx3} chooses the latter Tatar translation after the word рухани \eng{soul}.
+
+\begin{figure}
+\begin{lstlisting}
+  <rule>
+    <match lemma="рухани" tags="adj"/>
+    <match lemma="өмір" tags="n.*">
+        <select lemma="тормыш"/>
+    </match>
+  </rule>
+\end{lstlisting}
+\caption{A lexical selection rule that selects Tatar тормыш for Kazakh өмір when preceded by the word рухани.}
+\label{fig:lrx3}
+\end{figure}
+
+The system currently has a total of 33 lexical selection rules.
 
@@
 A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.
+Aside from the irregular behaviour of йөр, which would need its irregularities implemented through special phonotactics in lexc, these issues are all shortcomings of our phonological layer, implemented in twol.
 
-\begin{figure}[htbp]
+\begin{table}[htbp]
+	\centering
 	\begin{tabular}{lll}
@@
 	\label{tab:phonerrors}
-\end{figure}
-
-\todo{Example of lack of phon rule}
+\end{table}
 
@@
   \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
+  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \bottomrule
-  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \end{tabular}
-    \caption{Word error rate over two corpora. The \texttt{devel} `development' corpus was used during development 
+    \caption{Word error rate over two corpora. The development corpus (\texttt{devel}) was used during development 
        to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
@@
 
-%The majority of errors are currently due to mistakes and gaps in the morphophonology component; some minor problems still remain involving:
+%The majority of errors currently are due to one of the following:
 %\begin{itemise}
-%  \item Combinations of case and possessive suffixes,
-%  \item Orthographical representations of phonology,
-%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
+%   \item mistakes and gaps in the Tatar morphophonology component,
+%   \item lack of transfer rules to hande some of the Kazakh compund tense forms, and
+%   \item disambiguation errors.
+%%  \item Combinations of case and possessive suffixes,
+%%  \item Orthographical representations of phonology,
+%%  \item Vowel harmony processing on clitics (e.g., \emph{да}/\emph{дә} `and') after unknown words.
 %\end{itemise}
 
+The majority of errors currently are due to one of the following: mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
 %\cite{TyersAlperen2010}

### 2013-04-23 paper/paper.tex
submitted versions (full and anonymised)

@@
-\documentclass[11pt]{article}
+\documentclass[a4paper,11pt]{article}
 
@@
 
+% for anonymising
+% make a symlink to this file containing "anon" in the filename
+% then put anything to be deleted in that version in \anonymise{}
+% compile and enjoy
+\usepackage{ifthen}
+\usepackage{ifjobname}
+\ifjobinfix{anon}{
+	\newcommand{\anonymise}[1]{}
+}{
+	\newcommand{\anonymise}[1]{#1}
+}
+
 %%\Keywords{Kazakh, Tatar, MT, Free software, Open-source}
@@
 
-\author{Ilnar Salimzyanov\\
+\author{\anonymise{Ilnar Salimzyanov\\
   Kazan Federal University\\
@@
   E-03877 Alacant\\
-  \email{ftyers@dlsi.ua.es}}
+  \email{ftyers@dlsi.ua.es}}}
 
@@
 
-Within the Apertium project, work on MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Tatar--Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.
-Among these systems is a prototype Tatar--Bashkir machine translation system which was built by the authors of this paper \citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
+Within the Apertium project, work on several MT systems between Turkic languages has been started (Turkish--Kyrgyz, Azeri--Turkish, Tatar--Bashkir), but the Kazakh--Tatar system described by the present study is the closest to production-ready of them.
+Among these systems is a prototype Tatar--Bashkir machine translation system \anonymise{which was built by the authors of this paper }\citep{tyerswashingtonsalimzyanbattalov12}; due to the closeness of these languages, it proved to provide high accuracy in its translations, but being a prototype system by design, had relatively low coverage.
 
@@
 language, sharing official status with Russian as an official language.
-Large groups of native speakers also exist in China, neighbouring
+Large communities of native speakers also exist in China, neighbouring
 Central-Eurasian republics, and Mongolia. The total number of speakers is at least 10 million people.
@@
 
-\begin{figure*}[htbp]
+\begin{figure*}[bhtp]
 \hspace{3cm}\parbox[t]{0.7\textwidth}{{\tt
@@
 %FIXME: Describe the gist of the rules, maybe with a few concrete examples.
-The grammar currently has 60 rules.  The goal of these rules is to select the correct analysis when there 
-are multiple morphological analyses of Kazakh forms.
-%FIXME: It might be good to put some numbers about the morphological ambiguity of Kazakh
 
-%FIXME: JNW needs to copy-edit this, and we should probably provide a concrete example to make it clearer too
-The output of the morphological analyser is highly ambiguous. Both Kazakh and Tatar have a lot of affixes which join the
-verb stem. Some of the grammars label them all gerunds, or participles, assuming that there are no finite forms.
+The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.
+The goal of the CG rules is to select the correct analysis when there are multiple analyses.
+As of r43664, the ambiguity was down to 1.4 analyses per form.
 
-We give to a VerbStem+ParticularAffix surface form maximal number of readings (syntactic roles) it can have, such as [verbal
-adjective], [finite form], [substantivised verbal adjective, i.e., if the word it determines was omitted], whereas many grammars
-do not distinguish these and label forms of verbs containing a particular affix just as a certain gerund or participle form.
+One reason for the still high level of morphological ambiguity is a series of affixes in both languages which can each be analysed as some combination of verbal nouns, verbal adjectives, substantitivised verbal adjectives (i.e., verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}
 
-This type of ambiguity is common between the two languages we deal with, and therefore may be passed from one language
-to another and most likely will not lead to translation errors. Disambiguating it, and, prior to that, having that kind of
-informaton about the syntactical role of a surface form is crucial for success e.g. of a Turkic to English system.
+Given the similarity of Kazakh and Tatar, this sort of ambiguity may often be passed from one language to the other and not lead to many translation errors.
+While disambiguating between these analyses would be crucial for e.g., a Turkic-to-English system, we have not yet put much effort into developing CG rules to deal with such ambiguity.
 
-This is the reason for a high ambiguity rate of the transducers.
+%This is the reason for a high ambiguity rate of the transducers.
 
@@
 
-For example (see figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
+For example (see Fig.~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
 meanings corresponding to its Tatar cognate корал; however, it is also used in the 
@@
 to Tatar чара.
-A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in figure~\ref{fig:lrx}.
+A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in Fig.~\ref{fig:lrx}.
 
@@
 
-Likewise, the Kazakh word топ has two translations into Tatar.  One translation is туп \eng{ball} (which can also 
-be доп in Kazakh), and the other translation is төркем \eng{group}.  The bilingual dictionary also has 
+Likewise, the Kazakh word топ can be translated to Tatar as туп \eng{ball} (sometimes
+доп in Kazakh), and as төркем \eng{group}.  The bilingual dictionary also has 
 the Russian word группа \eng{group}, which is used in Tatar, as an entry which may be translated 
@@
 
-Another lexical selection rule is shown in figure~\ref{fig:lrx2}.
+Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).
+The lexical selection rule provided in Fig.~\ref{fig:lrx3} chooses the latter Tatar translation after the adjective рухани \eng{spiritual}.
+
+%Another lexical selection rule is shown in figure~\ref{fig:lrx2}.
 This rule ensures that the Kazakh word рет \eng{row, time(s)} is translated as тапкыр \eng{time(s)} when it is used with a number; otherwise it translates as рәт \eng{row}.
-\begin{figure}
-\begin{lstlisting}
-  <rule>
-    <match tags="num"/>
-    <match lemma="рет" tags="n.*">
-        <select lemma="тапкыр"/>
-    </match>
-  </rule>
-\end{lstlisting}
-\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
-\label{fig:lrx2}
-\end{figure}
-
-Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).
-The lexical selection rule is provided in figure~\ref{fig:lrx3} chooses the latter Tatar translation after the word рухани \eng{soul}.
+%\begin{figure}
+%\begin{lstlisting}
+%  <rule>
+%    <match tags="num"/>
+%    <match lemma="рет" tags="n.*">
+%        <select lemma="тапкыр"/>
+%    </match>
+%  </rule>
+%\end{lstlisting}
+%\caption{A lexical selection rule that selects Tatar тапкыр for Kazakh рет when preceded by a number.}
+%\label{fig:lrx2}
+%\end{figure}
 
@@
 \end{lstlisting}
-\caption{A lexical selection rule that selects Tatar тормыш for Kazakh өмір when preceded by the word рухани.}
+\caption{A lexical selection rule that selects тормыш for өмір if preceded by the word рухани.}
 \label{fig:lrx3}
@@
 
+\begin{table}[htbp]
+	\centering
+	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
+		\begin{tabular}{llrrr}
+			\toprule
+			Corpus          & Tokens & Coverage & stdev\\
+			\midrule
+			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
+			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
+			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
+			\bottomrule
+		\end{tabular}
+	}\\
+	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
+		\begin{tabular}{llrrr}
+			\toprule
+			Corpus          & Tokens & Coverage & stdev\\
+			\midrule
+			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
+			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
+			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
+			\bottomrule
+		\end{tabular}
+	}
+
+	\caption{Naïve coverage of the Kazakh-Tatar system}
+	\label{table:trimmedcoverage}
+\end{table}
 \begin{table*}[htbp]
@@
 
-For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty's Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
+For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty (RFERL)'s Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
 
-For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} the New Testament in Tatar, and content from Radio Free Europe / Radio Liberty's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.
+For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} a translation of the New Testament, and content from RFERL's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.
 
@@
 %%\end{table}
-
 \begin{table}
-	\centering
-	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
-		\begin{tabular}{llrrr}
+  \begin{center}
+  \begin{tabular}{ccrrr}
   \toprule
-			Corpus          & Tokens & Coverage & stdev\\
+   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
   \midrule
-			RFERL 2010      & 3.2M   & 90.19\% & $\pm$ 0.23\% \\
-			RFERL 2012      & 2.9M   & 89.74\% & $\pm$ 0.59\% \\
-			Wikipedia       & 1.2M   & 80.75\% & $\pm$ 5.23\% \\
+  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
+  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
   \bottomrule
   \end{tabular}
-	}\\
-	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
-		\begin{tabular}{llrrr}
-			\toprule
-			Corpus          & Tokens & Coverage & stdev\\
-			\midrule
-			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
-			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
-			Wikipedia       & 128K   & 81.36\% & $\pm$ 1.48\% \\
-			\bottomrule
-		\end{tabular}
-	}
-
-	\caption{Naïve coverage of the Kazakh-Tatar system}
-	\label{table:trimmedcoverage}
+    \caption{Word error rate over two corpora; OOV is the number of out-of-vocabulary (unknown) words.}
+% The development corpus (\texttt{devel}) was used during development 
+%       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
+%       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.
+ %\vspace{-2.5em}
+    \label{table:wer-development}
+  \end{center}
 \end{table}
@@
 As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
-system\footnote{Note that the coverage of the vanilla transducers is somewhat higher.} over 
+system\footnote{The coverage of the vanilla transducers is slightly higher.} over 
 the news corpora approaches that of a broad-coverage MT system, with one word in ten unknown.
-For the Wikipedia corpus, the coverage 
-is substantially worse.  This is due to the fact that this corpus is ``dirtier'' --- 
-containing more orthographical errors, \emph{wiki} code, and repetitions, as well as to the 
-fact that it has contains many more proper nouns.
+The coverage over the Wikipedia corpus 
+is substantially worse, due to the fact that this corpus is ``dirtier'': it contains orthographical errors, \emph{wiki} code, repetitions, as well as quite a few proper nouns.
 
@@
 We had two small Kazakh corpora along with their postedited translations into Tatar to measure the WER.
-The first one (2,457 words total) was a concatenation of an article from Radio Free Europe / Radio Liberty's
+The first one (2,457 words total) was a concatenation of an article from RFERL's
 Kazakh-language service, an article from Wikipedia, and a simple story used for pedagogical purposes in a workshop
-on MT for the languages of Russia. In addition to postediting the translation, this corpus was run through the
-morphological transducer and its output was manually disambiguated. All the stems in these texts had been added to
-the system, and all the rules (Constraint Grammar, lexical selection and transfer) were based on this corpus.
-Hence we can call it the development corpus. It presents an upper bound on the current performance
-of the system. Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN
+on MT for the languages of Russia. In addition to postediting the translation, we ran this corpus through the
+morphological transducer and manually disambiguated its output.  All the stems in these texts were added to
+the system, and all the rules (CG, lexical selection, and transfer) were based on this corpus.
+This ``development'' corpus presents an upper bound on the current performance
+of the system. %Table~\ref{table:wer-development} presents the WER for it. % http://pastebin.com/TaaWxJzN
+The testing corpus, used solely for evaluation, was comprised of articles from RFERL's Kazakh-language service, and had a similar size (2,862 words) to the development corpus.
+Table~\ref{table:wer-development} presents the WER for both corpora. % http://pastebin.com/7tzGCgMX.
 
-The second corpus --- a collection of articles from Radio Free Europe / Radio Liberty's Kazakh-language service --- was used solely for evaluation.  It had a similar size --- 2,862 words.
-The WER results for this corpus are presented in table~\ref{table:wer-development}. % http://pastebin.com/7tzGCgMX.
-
-Some of the corrections that were made as part of postediting were were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.
+Some of the corrections that were made as part of postediting were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.
 A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.
@@
 
-\begin{table}[htbp]
+\begin{table}
 	\centering
@@
 
-\begin{table}
-  \begin{center}
-  \begin{tabular}{ccrrr}
-  \toprule
-   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
-  \midrule
-  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
-  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
-  \bottomrule
-  \end{tabular}
-    \caption{Word error rate over two corpora. The development corpus (\texttt{devel}) was used during development 
-       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
-       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.}
- %\vspace{-2.5em}
-    \label{table:wer-development}
-  \end{center}
-\end{table}
-
 %Table~\ref{table:wer} presents the Word Error Rate, an edit metric based on the Levenshtein 
@@
 
-\subsection{Error analysis}
+%\subsection{Error analysis}
 
@@
 
-The majority of errors currently are due to one of the following: mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
+The majority of remaining errors are mostly due to mistakes and gaps in the Tatar morphophonology component, lack of transfer rules to handle some Kazakh compounds, and disambiguation errors.
 %\cite{TyersAlperen2010}
@@
 
-To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has a near-to-production-level
-coverage, but rather a prototype-level number of rules. Though the impact of this relatively low number of rules on the
-quality of the translation (compare the WER results for the corpus 1 and corpus 2) looks very promising and suggests
-the a high-quality translation between morphologically-rich agglutinative languages is possible.
+To our knowledge we have presented the first ever MT system between Kazakh and Tatar. It has near-production-level
+coverage, but is rather prototype-level in terms of the number of rules.  Although the impact of this relatively low number of rules on the
+quality of translation is extensive (cf., the difference in WER between the development and testing corpora), the outlook is promising and the current results suggest
+that a high-quality translation between morphologically-rich agglutinative languages is possible.
 
 We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
-increase it to 95\% on the corpora we have, the main work will be improving the quality of the translation by adding
+increase it to 95\% on the corpora we have, the main work will be improving the quality of translation by adding
 more rules, starting with the Constraint Grammar module.

### 2013-06-25 paper/paper.tex
new style file and clarification on Tatar verb types

@@
 Use of lexc allows for straightforward definition of different word classes and subclasses.
-For example, Tatar (but not Kazakh) has two classes of verbs: one which takes a harmonised high vowel in the infinitive (the default), and one which take a harmonised low vowel in the infinitive.
-This was implemented in lexc with two similar continuation lexica for verbs: one pointing at a lexicon with an A-initial infinitive ending, and another pointing at a lexicon with an I-initial infinitive ending.
+For example, Tatar (but not Kazakh) has two classes of verbs: one which takes a harmonised high vowel in the infinitive (the default), and one which takes a harmonised low vowel in the infinitive.
+Class membership cannot be predicted based on any phonological criteria and is simply a lexical property of any given verb.
+This was implemented in lexc with two similar sets of continuation lexica for verbs: one pointing at a lexicon with an A-initial infinitive ending, and another pointing at a lexicon with an I-initial infinitive ending.
+These two sets of continuation lexica are otherwise the same.
 

### 2013-06-25 paper/paper.tex
some minor stuff: * 2nd -> write 2^(nd) or second * table 1 -> Table 1 * Fig. 1 -> Figure 1

@@
 Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a 
-four-way distinction in Kazakh's 2nd person system (both pronouns and agreement suffixes), where Tatar only 
+four-way distinction in Kazakh's 2\super{nd} person system (both pronouns and agreement suffixes), where Tatar only 
 has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], 
-whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in table~\ref{tab:2pprons}.
+whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in Table~\ref{tab:2pprons}.
 \begin{table}[htbp]
@@
 	%\begin{subtable}
-	\subfloat[Kazakh 2nd person pronouns]{
+	\subfloat[Kazakh 2\super{nd} person pronouns]{
 		\begin{tabular}{lll}
@@
 	}\\
-	\subfloat[Tatar 2nd person pronouns]{
+	\subfloat[Tatar 2\super{nd} person pronouns]{
 		\begin{tabular}{lll}
@@
 	}
-	\caption{The 2nd person pronoun systems of Kazakh and Tatar}
+	\caption{The 2\super{nd} person pronoun systems of Kazakh and Tatar}
 	\label{tab:2pprons}
@@
 which have a one-to-one correspondence with the pronouns in each language, resulting in several systems of 
-suffixes in each language that have the same set of distinctions as in the 2nd person pronoun systems.
+suffixes in each language that have the same set of distinctions as in the 2\super{nd} person pronoun systems.
 
 The past tense systems of Kazakh and Tatar have a many-to-many correspondence.  As shown 
-in table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh 
+in Table~\ref{tab:pasttense}, at a basic level, in the past tense, Kazakh 
 differentiates [±eyewitness]\footnote{``Eyewitness'' is a convenient term for this feature, though it 
@@
 The Apertium translation engine consists of a Unix-style \emph{pipeline} or
-\emph{assembly line} with the following modules (see Fig.~\ref{fig:modules}):  
+\emph{assembly line} with the following modules (see Figure~\ref{fig:modules}):  
 \begin{itemise}
@@
 
-Entries consist largely of one-to-one stem-to-stem correspondences with part of speech, but also include some entries with ambiguous translations (see e.g., Fig.~\ref{fig:bidix}).
+Entries consist largely of one-to-one stem-to-stem correspondences with part of speech, but also include some entries with ambiguous translations (see e.g., Figure~\ref{fig:bidix}).
 
@@
 
-For example (see Fig.~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
+For example (see Figure~\ref{fig:bidix}), Kazakh құрал can mean an instrument, device, tool, or even weapon, all 
 meanings corresponding to its Tatar cognate корал; however, it is also used in the 
@@
 to Tatar чара.
-A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in Fig.~\ref{fig:lrx}.
+A lexical selection rule that selects the translation чара when it occurs in a compound with ақпарат is written to ensure the correct translation; this rule is shown in Figure~\ref{fig:lrx}.
 
@@
 Tatar has separate words for \eng{[physical] life} (гомер) and \eng{life [as a human condition]} (тормыш), whereas Kazakh only has one word (өмір).
-The lexical selection rule provided in Fig.~\ref{fig:lrx3} chooses the latter Tatar translation after the adjective рухани \eng{spiritual}.
+The lexical selection rule provided in Figure~\ref{fig:lrx3} chooses the latter Tatar translation after the adjective рухани \eng{spiritual}.
 
@@
 
-As shown in table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
+As shown in Table~\ref{table:trimmedcoverage}, the naïve coverage of the Kazakh-Tatar MT 
 system\footnote{The coverage of the vanilla transducers is slightly higher.} over 
@@
 Some of the corrections that were made as part of postediting were not translation errors as such, but were instead due to the lack of morphophonological rules in the Tatar transducer.
-A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in table~\ref{tab:phonerrors}.
+A few examples, which include irregular verbs (ди and йөр) and are otherwise orthographical corner cases, are given in Table~\ref{tab:phonerrors}.
 Aside from the irregular behaviour of йөр, which would need its irregularities implemented through special phonotactics in lexc, these issues are all shortcomings of our phonological layer, implemented in twol.

### 2013-06-25 paper/paper.tex
more minor fixes: * based on context rules the -> based on context rules, the * structural transfer module which -> module, which * CG and Constraint Grammar forms used * Table 4 does not appear in running text.

@@
   morphological information. 
-\item A \emph{morphological disambiguator} (constraint grammar) which chooses, using linguistic rules
+\item A \emph{morphological disambiguator} (CG) which chooses, using linguistic rules
   the most adequate sequence of morphological analyses for an ambiguous sentence. 
@@
 \item A \emph{lexical selection} module \citep{tyers12a} which chooses, based on context 
-  rules the most adequate translation of ambiguous source language LFs.
-\item A \emph{structural transfer} module which
+  rules, the most adequate translation of ambiguous source language LFs.
+\item A \emph{structural transfer} module, which
     performs local syntactic operations, is compiled from XML files containing rules that 
@@
 \end{itemise}
+
+Table~\ref{tab:pipeline} provides an example of a single phrase as it moves through the pipeline.
 
@@
  \caption{Translation process for the Kazakh phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
+ \label{tab:pipeline}
  %\vspace{-1em}
@@
 increase it to 95\% on the corpora we have, the main work will be improving the quality of translation by adding
-more rules, starting with the Constraint Grammar module.
+more rules, starting with the CG module.
 The long-term plan is to integrate the data created with other open-source data for Turkic languages in order to make

### 2013-06-25 paper/paper.tex
minor fixes to orthogrpahy section

@@
 Kazakh does not orthographically distinguish high unrounded vowels (/{\qipa ɘ}/ ‹і› and /ə/ ‹ы›) before 
-glides (/w/ ‹у› and /j/ ‹й›) by writing the combination with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both 
+glides (/w/ ‹у› and /j/ ‹й›) and represents both combinations with one letter; i.e., /{\qipa ɘ}j/ and /əj/ are both 
 written ‹и›, while /{\qipa ɘ}w/ and /əw/ are both written ‹у›.  The quality of these vowels is necessary to know in 
@@
 Additionally, Tatar and Kazakh both use `yoticised' vowels---i.e., when ‹о›, ‹у›, or ‹а› (along with ‹ә› in 
-Tatar) follow /j/, a single character is used to represent both: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,
+Tatar) follow /j/, a single character is used to represent the combination: ‹ё›, ‹ю›, and ‹я› respectively.\footnote{Furthermore, in Tatar, /j/ followed by ‹э› or ‹ы› in Tatar is represented by ‹е›, though ‹е› is also the non-word-initial variant of ‹э›.}  %Combined with the issues regarding ‹у› in Kazakh,
 

### 2013-06-30 paper/paper.tex
some cramming; still not quite crammed enough

@@
 
-\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\a\label{ex:like-kaz}
+%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\a\label{ex:like-kaz}
+%\begingl
+%\gla Маған сөздік түзген ұнайды. //
+%\glb маған сөздік түз-GAн ұна-E-дI //
+%\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
+%\glft \eng{I like writing dictionaries.} //
+%\endgl
+%\a\label{ex:like-tat}
+%\begingl
+%\gla Мин сүзлек төзергә яратам. //
+%\glb мин сүзлек төз-IргA ярат-E-м //
+%\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
+%\glft \eng{I like writing dictionaries.} //
+%\endgl
+%\xe
+
+\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\a
+\label{ex:like-kaz}
 \begingl
@@
 \glb маған сөздік түз-GAн ұна-E-дI //
-\glb \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
+\glc \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
 \glft \eng{I like writing dictionaries.} //
 \endgl
-\a\label{ex:like-tat}
+\xe
+%\a\label{ex:like-tat}
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\label{ex:like-tat}
 \begingl
@@
 \glb мин сүзлек төз-IргA ярат-E-м //
-\glb \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
+\glc \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
 \glft \eng{I like writing dictionaries.} //
@@
 
+%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\label{ex:kerek}
+%\a
+%\begingl
+%\gla Мен үйге қайтуым керек. //
+%\glb мен үй-GA қайт-у-Iм керек //
+%\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
+%\glft \eng{I need to go home.} //
+%\endgl
+%\a
+%\begingl
+%\gla Миңа өйгә кайтырга кирәк. //
+%\glb мин-GA өй-GA кайт-IргA кирәк //
+%\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
+%\glft \eng{I need to go home.} //
+%\endgl
+%\xe
+
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
 \label{ex:kerek}
-\a
+%\a
 \begingl
@@
 \endgl
-\a
+\xe
+
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\a
 \begingl
@@
 
-\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\label{ex:try}
+%\a
+%\begingl
+%\gla Айгүл оны табуға әрекет жасап жүр. //
+%\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
+%\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
+%\glft \eng{Aygül is trying to find him.} //
+%\endgl
+%\a
+%\begingl
+%\gla Айгөл аны табарга тырыша. //
+%\glb Айгөл а-NI тап-AргA тырыш-E //
+%\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
+%\glft \eng{Aygöl is trying to find him.} //
+%\endgl
+%\xe
+
+\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
 \label{ex:try}
-\a
+%\a
 \begingl
@@
 \glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
-\glb Aygül \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
+\glc Aygül he-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
 \glft \eng{Aygül is trying to find him.} //
 \endgl
-\a
+\xe
+\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\a
 \begingl
@@
 \label{ex:GAlI}
-\a
+%\a
 \begingl
@@
 \endgl
+\xe
 %\a
@@
 %\endgl
-\a
+\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+%\a
 \begingl
@@
 \glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
-\glb I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
+\glc I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
 \glft \eng{I came to speak with you.} //
@@
 }}%\end{texttt}
+\vspace{-0.5em}
 \caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
@@
 \end{lstlisting}
+\vspace{-1em}
 %}\\
@@
 %}
-\todo{A couple more lexical-selection rules }
 %\caption{Some lexical selection rules in Kazakh}
@@
 \end{lstlisting}
+\vspace{-1em}
 \caption{A lexical selection rule that selects тормыш for өмір if preceded by the word рухани.}
@@
 %{\bf Mor. disambiguation}& \^{}Һава\tag{<n>}\tag{<nom>\$} \^{}бүген\tag{<adv>\$} \^{}бик\tag{<adv>\$} \^{}әйбәт\tag{<adj>\$}\^{},\tag{<cm>\$} \\ 
-{\bf Mor.\ disambiguation} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}\tag{\$} \^{}ол\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}\tag{\$} \\
+{\bf Mor.\ disambig.} & \^{}Ол\tag{<prn>}\tag{<pers>}\tag{<p3>}\tag{<sg>}\tag{<nom>}\tag{\$} \^{}енді\tag{<adv>}\tag{\$} \^{}ол\tag{<det>}\tag{<dem>}\tag{\$} \^{}дыбыс\tag{<n>}\tag{<acc>}\tag{\$} \\
                      & \^{}анық\tag{<adj>}\tag{<comp>}\tag{<advl>}\tag{\$} \^{}есті\tag{<v>}\tag{<tv>}\tag{<prc\_impf>}\tag{\$} \^{}баста\tag{<vaux>}\tag{<ifi>}\tag{<p3>}\tag{<sg>}\tag{\$}\^{}.\tag{<sent>}\tag{\$} \\

### 2013-06-30 paper/paper.tex
more compressing

@@
         \end{itemize}
-        \vspace{-1pt}
+        \vspace{-2pt}
 }
@@
 			\toprule
-			 & {[$-$pl]} & {[$+$pl]} \\\midrule
-			{[$-$formal]} & сен & сендер \\
-			{[$+$formal]} & сіз & сіздер \\\bottomrule
+			 & {[-pl]} & {[+pl]} \\\midrule
+			{[-frml]} & сен & сендер \\
+			{[+frml]} & сіз & сіздер \\\bottomrule
 		\end{tabular}
@@
 	%}\subtable[Tatar 2nd person pronouns]{
-	}\\
+	}
 	\subfloat[Tatar 2\super{nd} person pronouns]{
@@
 			\toprule
-			 & {[$-$pl]} & {[$+$pl]} \\\midrule
-			{[$-$formal]} & син & сез \\
-			{[$+$formal]} & сез & сез \\\bottomrule
+			 & {[-pl]} & {[+pl]} \\\midrule
+			{[-frml]} & син & сез \\
+			{[+frml]} & сез & сез \\\bottomrule
 		\end{tabular}%}
@@
 	}
-	\caption{The 2\super{nd} person pronoun systems of Kazakh and Tatar}
+	\caption{2\super{nd} pers.\ pronoun systems of Kazakh and Tatar}
 	\label{tab:2pprons}
@@
 
-Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh 
+The morphotactics of the cognate Kazakh 
 distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
-different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although 
-in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in 
+different.\footnote{This comparison is made without regard to semantic alignment.}  Specifically, in both languages, the person agreement takes the form of a person copula suffix; however
+in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``you apparently went''), whereas in 
 Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
@@
 \pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 %\a
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 \label{ex:like-tat}
@@
 
-In Kazakh, a gerund (i.e., verbal noun) with case marking and sometimes person agreement in the form of 
-possessive suffixes is used to make a verb phrase an argument to certain other main phrases.  In Tatar, many 
-of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
+%In Kazakh, a gerund (i.e.,\ verbal noun) with case marking and sometimes person agreement (in the form of 
+%possessive suffixes) is used to make a verb phrase an argument to certain other main phrases.  In Tatar, many 
+%of these phrases use an invariant infinitival form.  Some examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
+%% Short version:
+In Kazakh, a gerund (or verbal noun), often with case marking and person agreement (via possessive suffixes) is used to make a verb phrase an argument to another main phrases.
+In Tatar, many of these phrases use an invariant infinitive form.
+Examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
 
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 \label{ex:kerek}
@@
 \xe
-
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 %\a
@@
 \pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 \label{ex:try}
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 %\a
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
+\vspace{-1ex}
 \label{ex:GAlI}
@@
 \xe
+
 %\a
@@
 \vspace{-0.5em}
-\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right}
+\caption{Example entries from the bilingual transfer lexicon. Kazakh is on the left, and Tatar on the right.}
 \label{fig:bidix}

### 2013-06-30 paper/paper.tex
only one orphan citation
down to 8 pages
monospace as spectie suggested :(

@@
 %\setlength{\bibspacing}{\baselineskip-0.4em}
+\usepackage{enumitem}
 
 \newenvironment{itemise}[1]{
-        \begin{itemize}\setlength{\itemsep}{-0.3em}
+        \begin{itemize}[itemsep=-0.3em,leftmargin=1em,itemindent=0em]
+		  %\setlength{\itemsep}{-0.3em}
+		  %%\setlength{\itemindent}{-1em}\setlength{\leftmargin}{0em}
 		  %\setlength{\itemindent}{-1em}
@@
 			\toprule
-			 & {[-pl]} & {[+pl]} \\\midrule
-			{[-frml]} & сен & сендер \\
-			{[+frml]} & сіз & сіздер \\\bottomrule
+			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
+			{\texttt{[-frml]}} & сен & сендер \\
+			{\texttt{[+frml]}} & сіз & сіздер \\\bottomrule
 		\end{tabular}
@@
 			\toprule
-			 & {[-pl]} & {[+pl]} \\\midrule
-			{[-frml]} & син & сез \\
-			{[+frml]} & сез & сез \\\bottomrule
+			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
+			{\texttt{[-frml]}} & син & сез \\
+			{\texttt{[+frml]}} & сез & сез \\\bottomrule
 		\end{tabular}%}
@@
 \end{table}
-
+%\vspace{-1ex}
 The morphotactics of the cognate Kazakh 
@@
 the grammatical subject and the infinitival verb phrase is the grammatical direct object.
+%\vspace{-1ex}
 %(does this count as syntactic?)
@@
 Examples are shown in (\ref{ex:kerek}-\ref{ex:try}).
-
+\vspace{-1ex}
 %\begin{exe}
@@
 
+\vspace{1ex}
 As shown in (\ref{ex:GAlI}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.
+%\vspace{-1ex}
 
@@
 %	\subfloat[A lexical selection rule that selects чара as the translation of құрал if part of a compound with ақпарат.]{
+\vspace{-1em}
 \begin{lstlisting}
@@
 The lexical selection rule provided in Figure~\ref{fig:lrx3} chooses the latter Tatar translation after the adjective рухани \eng{spiritual}.
+The system currently has a total of 33 lexical selection rules.
 
@@
 \begin{figure}
+\vspace{-1em}
 \begin{lstlisting}
@@
 
-The system currently has a total of 33 lexical selection rules.
+%The system currently has a total of 33 lexical selection rules.
 
@@
 			\midrule
-			RFERL 2007-2012 & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
+			RFERL 2007-'12   & 1.2M   & 82.24\% & $\pm$ 2.88\% \\
 			New Testament   & 137K   & 91.79\% & $\pm$ 1.39\% \\
@@
 \end{table}
+\begin{table}
+  \begin{center}
+  \begin{tabular}{ccrrr}
+  \toprule
+   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
+  \midrule
+  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
+  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
+  \bottomrule
+  \end{tabular}
+    \caption{Word error rate over two corpora; OOV is the number of out-of-vocabulary (unknown) words.}
+% The development corpus (\texttt{devel}) was used during development 
+%       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
+%       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.
+ %\vspace{-2.5em}
+    \label{table:wer-development}
+  \end{center}
+\end{table}
+
 \begin{table*}[htbp]
@@
 \end{tabular}
- \caption{Translation process for the Kazakh phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He begins to listen to that sound more carefully}.}
+ \caption{Translation process for the Kazakh phrase \emph{Ол енді ол дыбысты анығырақ ести бастады} \eng{He began to listen to that sound more carefully}.}
  \label{tab:pipeline}
@@
 %%\end{table}
-\begin{table}
-  \begin{center}
-  \begin{tabular}{ccrrr}
-  \toprule
-   Corpus                 & Direction           & Tokens  & OOV & WER (\%) \\
-  \midrule
-  \texttt{devel} & kaz$\rightarrow$tat & 2457    & 2       & 15.19 \\
-  \texttt{test} & kaz$\rightarrow$tat & 2862    & 43      & 36.57 \\
-  \bottomrule
-  \end{tabular}
-    \caption{Word error rate over two corpora; OOV is the number of out-of-vocabulary (unknown) words.}
-% The development corpus (\texttt{devel}) was used during development 
-%       to identify possible disambiguation, lexical-selection and transfer rules. The \texttt{test} corpus
-%       was not seen during development. The OOV column gives the number of out-of-vocabulary (unknown) words.
- %\vspace{-2.5em}
-    \label{table:wer-development}
-  \end{center}
-\end{table}
 

### 2013-06-30 paper/paper.tex
still too wide...

@@
 			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
-			{\texttt{[-frml]}} & сен & сендер \\
-			{\texttt{[+frml]}} & сіз & сіздер \\\bottomrule
+			{\texttt{[-frm]}} & сен & сендер \\
+			{\texttt{[+frm]}} & сіз & сіздер \\\bottomrule
 		\end{tabular}
@@
 			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
-			{\texttt{[-frml]}} & син & сез \\
-			{\texttt{[+frml]}} & сез & сез \\\bottomrule
+			{\texttt{[-frm]}} & син & сез \\
+			{\texttt{[+frm]}} & сез & сез \\\bottomrule
 		\end{tabular}%}

### 2013-06-30 paper/paper.tex
bleh.  well, the thing fits now

@@
 			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
-			{\texttt{[-frm]}} & сен & сендер \\
-			{\texttt{[+frm]}} & сіз & сіздер \\\bottomrule
+			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & сен\hspace{-4pt} & сендер\hspace{-4pt} \\
+			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сіз\hspace{-4pt} & сіздер\hspace{-4pt} \\\bottomrule
 		\end{tabular}
@@
 			\toprule
-			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
-			{\texttt{[-frm]}} & син & сез \\
-			{\texttt{[+frm]}} & сез & сез \\\bottomrule
+			 & {\texttt{[-pl]}} & {\texttt{[+pl]}}\hspace{-4pt} \\\midrule
+			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & син\hspace{-4pt} & сез \\
+			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сез\hspace{-4pt} & сез \\\bottomrule
 		\end{tabular}%}

### 2013-06-30 paper/paper.tex
explanation

@@
 
+An MT system between Tatar and Kazakh is potentially of great use to the language communities.
+Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).
+Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can be repurposed.
+
 %FIXME: Contrastive grammar
@@
 whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in Table~\ref{tab:2pprons}.
+\vspace{-1em}
 \begin{table}[htbp]
@@
 %\end{subtables}
+\vspace{-1ex}
 
@@
 \end{table}
+\vspace{-1ex}
 %\vspace{-1ex}
@@
 \pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\vspace{-1ex}
+\vspace{-1.5ex}
 %\a
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\vspace{-1ex}
+\vspace{-1.5ex}
 \label{ex:like-tat}
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\vspace{-1ex}
+\vspace{-1.5ex}
 \label{ex:kerek}
@@
 \pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\vspace{-1ex}
+\vspace{-1.5ex}
 %\a
@@
 \pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
-\vspace{-1ex}
+\vspace{-1.5ex}
 \label{ex:try}
@@
 
-\vspace{1ex}
-As shown in (\ref{ex:GAlI}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.
+%\vspace{1ex}
 %\vspace{-1ex}
+\vspace{0.8ex}
+
+As shown in (\ref{ex:GAlI}-\ref{ex:GAlI2}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.
+
+\vspace{-1ex}
 
@@
 \vspace{-1ex}
+\label{ex:GAlI2}
 %\a
@@
 The goal of the CG rules is to select the correct analysis when there are multiple analyses.
-As of r43664, the ambiguity was down to 1.4 analyses per form.
+Currently, ambiguity is down to 1.4 analyses per form.
+%As of r43664, the ambiguity was down to 1.4 analyses per form.
 
@@
 
+%FIXME NOW
+All evaluation was tested against version 0.1.0, or {\tt {\small rNNNNNN}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
+%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
+
 Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar.
@@
 
-The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
+%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
+Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.
 
@@
 The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
-from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
+from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
 

### 2013-06-30 paper/paper.tex
version submitted as final, though we still have a few things left to work on

@@
 %FIXME NOW
-All evaluation was tested against version 0.1.0, or {\tt {\small rNNNNNN}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
+All evaluation was tested against version 0.1.0, or {\tt {\small r45441}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
 %The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
@@
 The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
-from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
+from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
 

### 2013-07-01 paper/paper.tex
actually this is what we submitted
list of minor fixes ''The contents of Sections 2 and 3 do not correspond to the description given in Section 1.''

@@
 
-The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a brief
-description of the two languages; Section~\ref{sec:lang}\ gives a short review
-of some previous work in the area of Turkic--Turkic language translation and compares the grammar of Kazakh and Tatar;
+The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a short review of some previous work in the area of Turkic--Turkic language machine translation; Section~\ref{sec:lang} introduces Kazakh and Tatar and compares their grammar;
 Section~\ref{sec:sys}\ describes the system and the tools used to construct it;
@@
 Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).
-Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can be repurposed.
+Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can also be repurposed.
 

### 2021-05-27 paper/paper.tex
OFTC, http -> https, url fixups

@@
 system between Kazakh and Tatar.
-
 The paper will be laid out as follows: Section~\ref{sec:prev}\ gives a short review of some previous work in the area of Turkic--Turkic language machine translation; Section~\ref{sec:lang} introduces Kazakh and Tatar and compares their grammar;
@@
 Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).
-Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can also be repurposed.
+Some accompanying language resources (e.g.,\ morphololgical analysers, disambiguators) can also be repurposed.
 
@@
 potentially unreliable information and newly discovered information) and [±recent], whereas Tatar has only 
-three categories: eyewitness, non-eyewitness, and newly-acquired-information---all with no [±recent] distinction.  As 
+three categories: eyewitness, non-eyewitness, and newly-acquired information---all with no [±recent] distinction.  As 
 an example of the many-to-many correspondence that this results in, Tatar has a single non-eyewitness 
@@
 forms in Kazakh are used for both potentially unreliable information and newly acquired information, whereas 
-in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired-information (\texttt{-GAн- икән}) past forms are distinguished.
+in Tatar, non-eyewitness (\texttt{-GAн-}) and newly-acquired information (\texttt{-GAн- икән}) past forms are distinguished.
 
@@
 			\toprule
-			 & [$+$recent] & [$-$recent] \\\midrule
-			{}[$+$reliable] & -DI- & -GAн- \\
-			{}[$-$reliable] & -Iп- & -GAн екен- \\ \bottomrule
+			 & \texttt{[+recent]} & \texttt{[-recent]} \\\midrule
+			{}\texttt{[+reliable]} & -DI- & -GAн- \\
+			{}\texttt{[-reliable]} & -Iп- & -GAн екен- \\ \bottomrule
 		\end{tabular}
@@
 			\toprule
-			 & [$-$new] & [$+$new] \\\midrule
-			{}[$+$reliable] & -DI- & -GAн- икән \\
-			{}[$-$reliable] & -GAн- & --- \\\bottomrule
+			 & \texttt{[-new]} & \texttt{[+new]} \\\midrule
+			{}\texttt{[+reliable]} & -DI- & -GAн- икән \\
+			{}\texttt{[-reliable]} & -GAн- & --- \\\bottomrule
 		\end{tabular}
@@
 % * Мен оқуды жақсы көремін > Мин укырга яратам.
-% Also, see this page: http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
+% Also, see this page: https://wiki.apertium.org/wiki/Kazakh_and_Tatar/Pending_tests
 There are a number of minor syntactic differences in Tatar and Kazakh, which include differences in verb valencies in equivalent translations, as well Tatar's reliance on a ``true'' infinitive that is used in place of various verbal noun and verb adverb forms in Kazakh.
@@
 The system is based on the Apertium machine translation 
-platform \citep{apertium/2011}.\footnote{\url{http://www.apertium.org}} The 
+platform \citep{apertium/2011}.\footnote{\url{http://apertium.org}} The 
 platform was originally aimed at the Romance languages of the Iberian peninsula, but has also been adapted for 
@@
 
-The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.
+The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh and 2.1 for Tatar.
 The goal of the CG rules is to select the correct analysis when there are multiple analyses.
-Currently, ambiguity is down to 1.4 analyses per form.
+Currently, ambiguity is down to 1.4 analyses per form for Kazakh and 1.7 for Tatar.\footnote{The Tatar CG is largely based on rules generalised from Kazakh's CG, but it has not received as much attention.}
 %As of r43664, the ambiguity was down to 1.4 analyses per form.
 
-One reason for the still high level of morphological ambiguity is a series of affixes in both languages which can each be analysed as some combination of verbal nouns, verbal adjectives, substantitivised verbal adjectives (i.e., verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}
+One reason for the still high level of ambiguity is a series of affixes in both languages which can each be analysed variously as verbal nouns, verbal adjectives, substantitivised verbal adjectives (verbal adjectives with a null modified noun), and even finite forms.\footnote{Despite the fact that the various suffixes in this category pattern differently and do not form a single natural class, most grammars of the languages label all them as simply ``gerunds'' or ``participles''.}
 
@@
 	\subfloat[Naïve coverage of the Kazakh-Tatar direction]{
-		\begin{tabular}{llrrr}
+		\begin{tabular}{lrrr}
 			\toprule
@@
 	\subfloat[Naïve coverage of the Tatar-Kazakh direction]{
-		\begin{tabular}{llrrr}
+		\begin{tabular}{lrrr}
 			\toprule
@@
 %FIXME NOW
-All evaluation was tested against version 0.1.0, or {\tt {\small r45441}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
+All evaluation was tested against version 0.2.0, or {\tt {\small r45482}} in the Apertium SVN.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-kaz-tat}}
 %The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
@@
 Lexical coverage of the system is calculated over freely available corpora of Kazakh and Tatar.
-
 For Kazakh, two years worth of content (2010 and 2012) from Radio Free Europe / Radio Liberty (RFERL)'s Kazakh-language service,\footnote{\url{http://www.azattyq.org/}} as well as a recent dump of Wikipedia's articles in Kazakh\footnote{\url{http://kk.wikipedia.org/}; {\smallertt kkwiki-20130408-pages-articles.xml.bz2}} were used.
-
 For Tatar, a dump of articles from the Tatar Wikipedia,\footnote{\url{http://tt.wikipedia.org/}; {\smallertt ttwiki-20130205-pages-articles.xml.bz2}} a translation of the New Testament, and content from RFERL's Tatar-language service\footnote{\url{http://www.azatlyk.org/}} from early 2007 to early 2012 were used for testing.
@@
 
-We plan to continue development on the pair; the coverage of the system is already quite high, and although we intend to
-increase it to 95\% on the corpora we have, the main work will be improving the quality of translation by adding
+We plan to continue development on the pair; the coverage of the system is already quite high, although we intend to
+increase it to 95\% on the corpora we have --- we estimate that this will mean adding around 5,000 new stems and take 1--2 months.
+The remaining work will be improving the quality of translation by adding
 more rules, starting with the CG module.
@@
 The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
-from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
+from sourceforge.\footnote{\url{http://sourceforge.net/projects/apertium/files/apertium-kaz-tat/}}
 

### 2021-05-27 dev/trimmed-coverage.sh

@@ -9,7 +9,7 @@ fi
 # This file is kept even after the script is ended:
 needed=/tmp/corpus-stat-all-needed.txt
 
-covgoal="99.9" # percent goal we aim for this week, according to http://wiki.apertium.org/wiki/Kazakh_and_Tatar/Work_plan
+covgoal="99.9" # percent goal we aim for this week, according to https://wiki.apertium.org/wiki/Kazakh_and_Tatar/Work_plan
 
 
 cd "$(dirname $0)"
